---
title: "MRManalyzeR: Reproducible QC and stats reporting for targeted metabolomics"
author:
- name: Matthew J. Smith
  affiliation: Karolinska Institutet, Stockholm, Sweden
  email: mattyjsmith123@gmail.com
- name: Antonio Checa
  affiliation: Karolinska Institutet, Stockholm, Sweden
date: "`r Sys.Date()`"
package: MRManalyzeR
output:
  BiocStyle::html_document:
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{MRManalyzeR: Reproducible QC and stats reporting for targeted metabolomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-knitr, include = FALSE}
knitr::opts_chunk$set(comment = "#>", fig.width = 6.5, fig.height = 4,
                      dpi = 100, message = FALSE, warning = FALSE)

# Tables are shown the way the HTML reports show them: searchable,
# sortable and scrollable rather than truncated with head().
dt <- function(x, rownames = TRUE, page = 5){
  DT::datatable(x, rownames = rownames, filter = "top",
                options = list(pageLength = page, scrollX = TRUE,
                               dom = "ftip"))
}
```

# Introduction

Targeted LC-MS/MS data processing is less complex than untargeted profiling,
though the assumptions differ. In untargeted profiling the analytical hurdle
comes after acquisition; deconvolution, identifying unknowns, and the
bioinformatics that follows. Targeted LC-MS/MS has the complexity upfront: the
compound panel is defined in advance, so what remains afterwards are questions
of **data quality** and the **accurate reporting of known metabolites** rather
than of feature discovery. Normalisation to stable-isotope internal standards is
commonplace, and researchers may report raw peak areas, internal-standard
adjusted relative ratios, or absolute concentrations depending on the study
design or quantitative goals.

MRManalyzeR is built specifically for targeted metabolomics data, focused on
assessing data quality across flexible quantitative outputs -from raw areas to
true concentrations- and giving a basic interpretation of the known analytes
that come out. The output is deliberately visual, where quality control and
statistics are presented as plots with interpretable statistical summaries
(e.g. coefficient of variation), not as a modelling framework. Everything is
driven from a curated `.xlsx` workbook and a single YAML file, so a run is
reproducible and the workflow stays accessible to analysts without a coding
background.

The YAML configuration doubles as documentation of the run. Inputs and
parameters are validated before any processing begins -- `validateInput()`
checks the workbook and `validateDesign()` the study design. Each step records
how the data was handled, from the number of features a blank filter removed to
the imputation and scaling a PCA applied. Quality control therefore rests on a
traceable record of how a reported number was arrived at, not on the final
figure alone -- in line with FAIR principles (Wilkinson *et al.*, 2016).

The package currently reads Waters **TargetLynx** exports (the primary format),
**Skyline** exports, and plain sample-by-analyte matrices from any other
software, and covers signal filtering, blank filtering, batch correction,
normalisation, concentration adjustment and missing-value imputation,
followed by quality control, statistics and two self-contained HTML reports.

## Scope, and where this sits among related packages

MRManalyzeR begins with **reviewed vendor result tables**. It does not perform
chromatographic peak detection, peak integration or calibration-curve fitting
from raw mass-spectrometry data; these decisions are left to the analyst's
domain expertise, using their preferred software (e.g. TargetLynx or Skyline) to
inspect and curate the underlying chromatograms.

That is the line which separates it from the Bioconductor packages nearest to
it, and from the existing targeted-MRM tools:

| Task | Tool |
|---|---|
| Raw untargeted LC-MS feature detection and alignment | `xcms` |
| General metabolomics peak-matrix preprocessing | `pmp` |
| Metabolite annotation-table workflows | `MetMashR` |
| Chromatogram processing, peak integration, automated identification | MRMAnalyzer, MRMkit, MRMQuant |
| Post-acquisition targeted-assay processing, QC and reporting | **MRManalyzeR** |

The design assumes a validated targeted panel of >20 MRM transitions, known
compounds carrying class or pathway annotation, and an experiment structured
around defined target lists -- whether testing a specific hypothesis or
conducting targeted exploratory profiling. Feature metadata is treated as
biology, not bookkeeping -- for example compound class and enzymatic pathway
groups, used to colour heatmap annotations, PCA loadings and correlation blocks
throughout.

## Assumptions and limitations

Worth knowing before you rely on the output.

* **Batch correction equalises batch medians.** It assumes the reference
  samples (`QC`, or biological samples if the design is balanced across
  batches) are comparable between batches. `correctBatch()` refuses to run
  without enough reference injections per batch, and warns when the batch
  variable is confounded with a study factor -- but it cannot detect a design
  where that confounding is total.
* **Normalisation and quantification depend on assay design.** MRManalyzeR
  accommodates a range of targeted workflows, whether reporting unadjusted peak
  areas or fully calibrated concentrations. Its built-in normalisation and
  batch-adjustment tools assume conservative corrections, however; higher-tier
  absolute quantification depends on the user's study design (e.g. validated
  methods with isotope-labelled standards). Where assays lack internal
  standards to absorb matrix variation and signal drift, packages such as `pmp`
  offer dedicated spline-based corrections (e.g. QC-RSC).
* **Concentration adjustment from provided values.** If needed, it applies the
  internal-standard concentration ratio and the sample-volume correction
  (assuming sample vial concentration is provided); it does not fit or refit a
  calibration curve.
* **Imputation draws values below a fixed fraction of each compound's observed
  minimum.** That is a convention that can be modulated based on the user's
  expertise, not an estimate. Values below the detection limit and values
  missing for other reasons are not currently distinguished.
* **Statistical methods are chosen by the user, per comparison.** Nothing is
  selected automatically from the data, and no test is run in parallel with
  another so that the more convenient answer can be picked afterwards.
* **Multiple-testing correction is applied within a family**, never across the
  whole workbook.

## How the package is organised

The workflow falls into four stages, preceded by a validation pass that checks
the workbook and the study design before any data is processed. Everything from
stage 1 onwards is carried in a single `struct::DatasetExperiment` -- the assay
matrix, the sample metadata and the feature metadata kept aligned in one object
-- so every function takes that object and returns it, and nothing has to be
handed a matching metadata table a second time.

| Stage | Question it answers | Functions |
|---|---|---|
| 0. Validation | Is the input usable, and the design sound? | `validateInput()`, `validateDesign()` |
| 1. Data parse | What did the instrument report? | `readTargetLynx()`, `readSkyline()`, `assembleDataset()`, `loadConfig()`, `loadDataset()`, `combineDatasets()` |
| 2. Peak-matrix processing | What is the true concentration? | `filterBlanks()`, `correctBatch()`, `normaliseMatrix()`, `adjustConcentration()`, `imputeMissing()`, `subsetDataset()`, `processDataset()` |
| 3. QC check | Can these numbers be trusted? | `addCVMetrics()`, `summariseFeatures()`, `runPCA()`, `plotDrift()`, `plotQQ()`, `plotPCA()`, `plotLoadings()` |
| 4. Stats | What do these numbers mean? | `runStats()`, `summariseGroups()`, `plotBoxplot()`, `plotVolcano()`, `plotHeatmap()`, `plotGroupHeatmap()`, `plotCorrelation()` |

```
   project_config.yml            workbook.xlsx
           |                           |
           |                  validateInput()              [0]
           |                  validateDesign()
           |                           |
           |                  readTargetLynx() / readSkyline()   [1]
           |                           |
           |                  wide sample x compound matrix
           |                           |
           |                    assembleDataset()
           |                           |
           |                           v
           |                   DatasetExperiment
           |                           |
           |          filterBlanks -> filterFeatures ->    [2]
           |          filterSamples -> correctBatch ->
           +--------> normaliseMatrix -> adjustConcentration
                      -> imputeMissing
                                       |
                                       v
                          processed DatasetExperiment
                               |          |          |
             [3] QC check      |          |          |   [4] Stats
             addCVMetrics      |          |          |   summariseGroups
             runPCA            |          |          |   runStats
             summariseFeatures |          |          |
                               |          |          |
             tables + plots    |          |          |tables + plots
                               v          v          v
             data_quality      .xlsx      .RDS       stats_report
               _report.html                          .html
                                         |
                                    _stats.xlsx
```

Every number in a figure comes from the same table that is written to the
spreadsheet -- `runStats()`, `summariseGroups()` and `addCVMetrics()`
compute once, the `plot_*()` functions consume. A figure and the sheet beside
it cannot disagree.

Each plot is an exported function rather than code buried in a report
template, and that is deliberate. Most users will never call one: the YAML
route renders both reports and the figures appear. But when a figure needs to
go into a paper with different labels, or one compound needs looking at on its
own, or a reviewer asks for the same plot on a subset, the choice should not be
between editing an R Markdown template and starting from the raw matrix.
`plotVolcano(st, "PBS_vs_HDM")` is one line, takes the objects the workflow
already produced, and returns an ordinary `ggplot` you can add layers to,
restyle or save. The reports call exactly the same functions, so what you get
interactively is what the report shows.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("MRManalyzeR")
```

The development version:

```{r install-dev, eval = FALSE}
BiocManager::install("MJS-708/MRManalyzeR")
```

```{r load}
library(MRManalyzeR)
```

# Quick start

`runExample()` runs all four stages on the bundled dataset and writes its
outputs to a temporary directory. `render = FALSE` builds the peak matrix and
the xlsx/RDS outputs but skips the (slower) HTML report rendering; drop it to
get the reports as well. Running this successfully is the recommended check
that an installation works before pointing the package at real data.

```{r run-example}
res <- runExample(render = FALSE, open = FALSE)
list.files(res$output_directory)
```

Everything after this section unpacks what that single call did.

# Stage 1: data parse

## The curated workbook

MRManalyzeR expects one `.xlsx` workbook holding the instrument export plus the
two curation sheets that describe it:

| Sheet | One row per | Purpose |
|---|---|---|
| `lcms_data` (or `skyline_data`) | peak | the instrument export itself |
| `feature_metadata` | compound | which compounds to report, and their annotation |
| `sample_metadata` | injection | which injections to include, and their design |

Keeping the curation next to the data is what makes a run reproducible: the
decisions about which compounds are reportable and which injections belong to
the study are recorded in the workbook, not in a script.

Two practical rules govern whether a run will work at all:

* **Names must match exactly.** `feature_metadata$Processing_name` has to match
  the compound names in the acquisition method, and `sample_metadata$Name` has
  to match the injection names in the data sheet -- including trailing spaces,
  which are the usual culprit.
* **Exclusions are recorded, not deleted.** An injection you do not want
  analysed gets `Include = "NO"` rather than being removed; a compound that
  should not be reported gets `Report = "NO"`. Re-injections are the common
  case, and marking them keeps the workbook a faithful record of the run.

```{r dataset}
xlsx  <- system.file("extdata", "example_data.xlsx",
                     package = "MRManalyzeR")
fdata <- openxlsx::read.xlsx(xlsx, sheet = "feature_metadata")
meta  <- openxlsx::read.xlsx(xlsx, sheet = "sample_metadata")

c(injections = nrow(meta), compounds = nrow(fdata))
```

The bundled workbook is a subset of the oxylipin LC-MS/MS data of Kolmert
*et al.* (2018) (see [References](#references)): 79 injections of mouse
bronchoalveolar lavage over a calibration ladder, extraction blanks, pooled QC
injections and 45 study samples.

```{r dataset-samples}
dt(as.data.frame(table(Sample_type = meta$Sample_type)),
   rownames = FALSE)
```

Because the panel is known in advance, `feature_metadata` can carry real
biology alongside the reporting decision -- the enzymatic pathway a compound
comes from, its precursor PUFA, its retention time. Those columns serve a
functional role: they colour the PCA loadings, group the heatmap rows and define
the metabolite ratios, and they travel with the data as `variable_meta` all the
way to the output.

```{r dataset-features}
dt(fdata[fdata$Report == "YES",
         c("Compound", "Enzymatic_pathway", "PUFA", "RT", "Report")],
   rownames = FALSE)
```

## The TargetLynx export

TargetLynx does not export a matrix. Its "complete summary" (`Edit > Copy >
Complete Summary`, pasted straight into the `lcms_data` sheet) is a *stacked*
table -- one block per compound, each block its own little table of injections,
with the compound name in a header row between blocks. It is tedious to inspect
visually and does not parse into R with existing tools, which is the first
problem the package solves.

`readTargetLynx()` parses those blocks, pivots them into a wide
**sample x compound** matrix of the requested `datatype` column, and (with
`snr`) masks any peak whose signal-to-noise falls below the threshold:

```{r read}
m <- readTargetLynx(xlsx, datatype = "Area", snr = 3)
dim(m)
```

```{r read-table}
dt(round(m[, 1:6], 1))
```

Columns come back as the raw TargetLynx compound identifiers
(`Processing_name`). `feature_metadata` maps those to the canonical `Compound`
name used everywhere downstream, which is how an instrument method can be
renamed without touching the analysis.

`readSkyline()` returns the same contract from a Skyline molecule x sample
sheet, with `signal_filter = "LOD"` or `"LOQ"` in place of the S/N mask.

Output from any other software can be supplied as a plain sample-by-analyte
matrix, which enters the workflow at `assembleDataset()` and is processed
identically from there on. Readers for further vendor formats are planned for a
later version; suitable test data is being sought, since a reader written
without real exports to check it against is not one worth relying on.

## The simulated dataset used from here on

The example above is a single chromatographic batch, injected from one
reconstitution volume, with no per-sample amount to normalise to so
normalisation, concentration adjustment and batch correction would all be
incompatible. The package therefore bundles a second, **entirely simulated**
workbook: a Skyline-layout export of 20 analytes across three enzymatic
pathways, 54 injections split over two chromatographic batches, with a
per-sample protein amount and a different reconstitution volume per batch.

The study design is balanced across the batches, so the batch effect and the
treatment effect are not confounded.

```{r syn-data}
sxlsx <- system.file("extdata", "example_synthetic.xlsx",
                     package = "MRManalyzeR")
sfeat <- openxlsx::read.xlsx(sxlsx, sheet = "feature_metadata")
smeta <- openxlsx::read.xlsx(sxlsx, sheet = "sample_metadata")

table(smeta$Chrom_Batch, smeta$Sample_type)
```

## Assembling a DatasetExperiment

`assembleDataset()` aligns the matrix with its feature and sample metadata and
returns a `struct::DatasetExperiment`. **Every step from here on takes and
returns that object**, so the metadata travels with the data and no step has to
be handed a matching sample table again.

```{r step-assemble}
sm <- readSkyline(sxlsx, "skyline_data", sfeat, signal_filter = "LOD")
de <- assembleDataset(sm, sfeat, smeta)
de
```

# Stage 2: peak-matrix processing

## Blank filter

`filterBlanks()` finds the extraction blanks via `sample_meta` and sets any
value below `mean(blanks) * blank_filter` to `NA`. A factor of 3 is a common
choice -- a peak has to be at least three times the blank level to be treated
as real.

```{r step-blanks}
de <- filterBlanks(de, blank_filter = 3, blank_head = "Sample_type",
                    blank_name = "Blank")
sum(is.na(de$data))
```

## Detection filters

A compound that was never really detected still occupies a column, and carries
into the statistics and the heatmap as a row of mostly-missing values - and into
an imputed workbook as mostly-invented ones. `filterFeatures()` removes those
and `filterSamples()` removes injections that missed most of the panel. Both
are off by default.

The subtlety is *which* samples the detection fraction is computed over. Take a
compound present throughout one treatment group and absent from the others --
often the most interesting thing in a dataset:

```{r step-filter-demo}
demo <- assembleDataset(
  data.frame(everywhere    = c(10, 12, 11, 9, 10, 11),
             group_specific = c(NA, NA, NA, 20, 22, 21),
             sparse         = c(5, NA, NA, NA, NA, 6),
             row.names = paste0("S", 1:6)),
  # assembleDataset() keeps only rows flagged Report == "YES"
  data.frame(Compound = c("everywhere", "group_specific", "sparse"),
             Report   = "YES"),
  data.frame(Name        = paste0("S", 1:6),
             Sample_type = rep("Sample", 6),
             Treatment   = rep(c("PBS", "HDM"), each = 3)))

across <- filterFeatures(demo, min_frac = 0.5, method = "across")
within <- filterFeatures(demo, min_frac = 0.5, method = "within",
                         group_col = "Treatment")

list(across = colnames(across$data), within = colnames(within$data))
```

`"across"` pools all six samples, so `group_specific` scores 3/6 and is
deleted *because* it is group-specific -- the classic way to lose the finding
you were looking for. `"within"` scores each treatment separately and keeps a
feature that clears the threshold in **any one** of them, so it survives on the
group it belongs to while `sparse` still goes. A third mode, `"QC"`, scores
against the QC injections and answers a different question -- whether a
compound is measured *reproducibly* rather than whether it is present -- and
falls back to `"within"` with a warning when a run has no QCs.

Small groups quantise the threshold: with `n = 3` a fraction can only be 0,
1/3, 2/3 or 1, so every `min_frac` between 1/3 and 2/3 behaves identically.

QC and blank injections never count towards `"within"` or `"across"`. They
stay in the dataset -- the blank filter and `correctBatch()` still need them --
but a compound seen only in the QCs cannot keep itself alive. The fraction each
feature was judged on is kept as `detected_frac`.

On the working dataset, where `readSkyline()` has already masked values below
the LOD:

```{r step-filter-features}
de <- filterFeatures(de, min_frac = 0.5, method = "within",
                     group_col = "Treatment")
```

`filterSamples()` then drops injections by missing-value fraction. Run it
*after* `filterFeatures()`: judged against a feature list that still contains
everything the instrument barely saw, every injection looks poor because the
denominator is padded with noise. Only study samples are removed -- a QC over
the threshold is reported but kept, since dropping one can leave a batch with
nothing for `correctBatch()` to anchor to.

```{r step-filter-samples}
de <- filterSamples(de, max_na = 0.8)
summary(as.data.frame(de$sample_meta)$na_frac)
```

Both run once over the whole dataset rather than per batch: a detection
fraction is a property of the study, and filtering per batch would leave each
batch with a different set of columns. In a YAML config they are the
`filter_features:` and `filter_samples:` blocks, applied after the blank filter
and before imputation -- imputation fills the very cells these measure.

## Batch correction

Where an acquisition spans several days or columns, response can shift between
batches. `correctBatch()` scales each batch so its per-feature median matches
the grand median across the reference samples. It runs straight after the
filters, on the instrument signal, and before normalisation, concentration
adjustment and imputation: the QC medians are then not rescaled by per-sample
amounts that pooled QCs usually do not have, and they come from what was
measured. A missing value counts as zero:
a compound missing from most of a batch's QCs is described by that, not by the
few injections where it was found. Where more than half are missing the median
is zero, and a compound that fails like that in any batch is corrected in
none; the failing batches are printed and kept in `bc_failed_batches`. To
correct only part of the panel, pass
`feature_col` -- a `feature_metadata` column of `YES` / `NO`.

```{r step-batch}
batch     <- stats::setNames(de$sample_meta$Chrom_Batch,
                             rownames(as.data.frame(de$data)))
de_uncorr <- de

de <- correctBatch(de, qc_label = "QC",
                    factor_name = "Sample_type",
                    batch_head  = "Chrom_Batch")
de_bc     <- de
```

Taking the most completely measured analyte as an example, the two batch
medians -- counting a missing value as zero, as the correction does -- are
pulled together:

```{r step-batch-medians}
j   <- which.min(colSums(is.na(as.data.frame(de_uncorr$data))))
med <- function(x) tapply(as.data.frame(x$data)[[j]], batch,
                          function(v) median(replace(v, is.na(v), 0)))

rbind(before = med(de_uncorr),
      after  = med(de))
```

A compound whose reference median is zero in any batch is left as measured in
all of them rather than corrected in some. In this dataset the two analytes
simulated near their detection limit are below it in more than half of the
batch-2 QCs, which puts that median at zero. They are listed here, and they
still carry the batch difference in the PCA further down:

```{r step-batch-failed}
vm <- as.data.frame(de_bc$variable_meta)
vm[nzchar(vm$bc_failed_batches), c("Compound", "bc_failed_batches")]
```

Use pooled QC injections as the reference (`qc_label = "QC"`) whenever the
design has enough of them in every batch; falling back to `"Sample"` assumes
the batches are biologically comparable.

## Normalise

`normaliseMatrix()` divides each sample by a known per-sample amount --
protein, tissue weight, or the volume of biofluid extracted -- so that
concentrations are expressed per unit of starting material:

```{r step-normalise}
de <- normaliseMatrix(de, column = "protein_ug")
round(as.data.frame(de$data)[1:3, 1:3], 3)
```

## Concentration adjustment

A calibration curve returns the concentration *in the vial the instrument
sampled*, and it does so based on the calibrant vial: the curve was built at the
calibrant's internal-standard concentration, and it assumes the sample carries
the same. `adjustConcentration()` corrects that assumption (based on the same
volume of calibration and study sample being injected into the instrument):

$$
\text{conc} \times \frac{\text{cal vol}}{\text{sample vol}}
            \times \frac{\text{sample IS}}{\text{cal IS}}
$$

Ultimately leading to:

$$
\frac{\text{cal vol}}{\text{sample vol}} \times
\frac{\text{sample IS}}{\text{cal IS}}
\;=\;
\frac{\text{sample IS} / \text{sample vol}}{\text{cal IS} / \text{cal vol}}
\;=\;
\frac{[\text{IS}] \text{ in the sample vial}}
     {[\text{IS}] \text{ in the calibrant vial}}
$$

So the correction is the ratio of internal-standard *concentrations*. More vial
volume per unit of IS means the IS is more dilute in your sample than in the
calibrant; a smaller IS peak inflates the analyte/IS response ratio, the curve
reads high, and this factor -- below 1 -- scales it back down.

Note what is *not* being corrected here. Reconstituting in more solvent dilutes
the analyte and the internal standard equally, so the response ratio does not
move and neither does the curve's answer.

Setting `starting_vol_col` adds a second step, back to the concentration in the
original, undried sample:

$$
\text{conc}_{\text{original}} = \text{conc}_{\text{vial}} \times
\frac{\text{sample vol}}{\text{starting vol}}
$$

```{r step-adjust}
de <- adjustConcentration(de,
                           sample_vol_col   = "sample_volume_uL",
                           cal_vol_col      = "cal_vol_uL",
                           sample_IS_col    = "sample_IS_vol_uL",
                           cal_IS_col       = "cal_IS_vol_uL",
                           starting_vol_col = "starting_vol_uL")
round(as.data.frame(de$data)[1:3, 1:3], 5)
```

## Impute missing values

`imputeMissing()` fills the remaining `NA`s in the biological samples with
values below that compound's lowest measurement, treating a non-detect as lying
under the detection floor rather than at zero. The ceiling is `scalar` times the
minimum -- 0.2 here -- and by default every cell gets its own random draw
between 0 and that ceiling, from a normal centred mid-range. A single constant
would give every non-detect the same value, which shrinks a compound's variance
and ties its ranks; `method = "constant"` does that if it is wanted. The draws
use a fixed `seed`, so a rerun is identical. Blanks are left untouched.

The PCA pre-processing (`runPCA(impute = )`) offers the same random fill
(`"gaussian"`, its default) alongside `"frac_min"`, `"min"`, `"half_min"` and
`"none"`, and missing-value filters that drop features or samples above a
missingness threshold instead of imputing them. For targeted panels a compound
missing in most samples is usually better dropped than modelled.

Imputation comes last, after batch correction, normalisation and concentration
adjustment, and it only feeds the exported matrix: in a YAML-driven run the
reports use the measured values, and their PCAs impute for themselves, unless
`use_imputed: True` is set.

```{r step-impute}
de <- imputeMissing(de, scalar = 0.2, blank_head = "Sample_type",
                     blank_name = "Blank")
sum(is.na(de$data))
```

Imputation is the one step that invents numbers, so it records what it did in
`variable_meta`: the ceiling (`impute_max`), how many cells it filled
(`n_imputed`) and what share of the compound's samples that is
(`frac_imputed`). A compound with a high `frac_imputed` is mostly invented, and
a difference found in it says more about the fill than about the biology:

```{r step-impute-record}
vm <- as.data.frame(de$variable_meta)
head(vm[order(-vm$frac_imputed),
        c("impute_max", "n_imputed", "frac_imputed")])
```

## The whole matrix in one call

`processDataset()` composes stages 1 and 2 -- read, assemble, filter,
batch-correct, normalise, adjust, then impute -- driven by the same metadata
sheets:

```{r step-process}
out <- processDataset(
  sfeat, smeta,
  xlsx_path        = sxlsx,
  data_source      = "skyline",
  data_tab_names   = "skyline_data",
  signal_filter    = "LOD",
  blank_filter     = 3,
  normalize        = "protein_ug",
  adjust_conc      = TRUE,
  starting_vol_col = "starting_vol_uL",
  replace_MVs      = 0.5,
  batch_correction = TRUE,
  bc_qc_label      = "QC",
  bc_header        = "Chrom_Batch",
  blank_head       = "Sample_type")

dim(out[[1]]$data)
```

`out[[2]]` holds the features dropped along the way with the reason recorded --
the same table the data-quality report surfaces as "compounds excluded", and
the first place to look when a compound you expected is missing from the
output.

`out$measured` is the same dataset before imputation, gaps included -- what
the reports and statistics use in a YAML-driven run unless `use_imputed: True`
is set. `out[[1]]` has its gaps filled everywhere except in the blanks, which
are left as they were:

```{r step-process-measured}
sum(is.na(out$measured$data))                         # gaps as measured
sum(as.data.frame(out[[1]]$variable_meta)$n_imputed)  # filled in out[[1]]
```

The processed dataset also ships ready-made, which is what the rest of this
vignette uses:

```{r syn-load}
syn <- loadDataset(system.file("extdata", "example_synthetic.RDS",
                                package = "MRManalyzeR"))
dim(syn$data)
```

# Stage 3: QC check

## How well was each compound measured?

`addCVMetrics()` writes the per-compound quality metrics into
`variable_meta`, matched to the feature they describe, so annotation and
quality live in the same table. `runMRManalyzeR()` applies it to every run.

```{r qc-cv}
syn <- addCVMetrics(syn, qc_label = "QC")

cv <- as.data.frame(syn$variable_meta)[
  , c("Compound", "Enzymatic_pathway", "CV_QC", "CV_sample",
      "CV_sample_vs_QC")]

dt(cv[order(cv$CV_QC), ], rownames = FALSE, page = 10)
```

* **`CV_QC`** is the coefficient of variation across the pooled QC injections.
  Every QC is the same material, so this is **technical** variability.
* **`CV_sample`** is the same statistic across the biological samples, so it
  carries the **biological** spread *plus* that same technical noise.
* **`CV_sample_vs_QC`** is `CV_sample / CV_QC`. A ratio of biological and
  technical variance, where higher values show most variance is biological.
  A ratio above 2 is a commonly cited reference point (Broadhurst *et al.*,
  2018), though the analyst should decide.

## Analytical drift

`plotDrift()` puts one compound's response against acquisition order, coloured
by sample type so the pooled QCs can be tracked through the sequence. A trend,
a step or a widening spread in the QCs is instrumental, not biological.

```{r qc-drift}
plotDrift(syn, "Analyte_02", value_label = "ng/mL")
```

## Normality

`plotQQ()` asks whether a compound's distribution supports the parametric
tests. Points along the line mean the t-test is on safe ground; a curve means
the Wilcoxon result reported alongside it is the one to trust. Targeted panels
are frequently right-skewed, which is why the report renders a tab per
transform rather than picking one.

```{r qc-qq, fig.width = 5, fig.height = 4}
plotQQ(syn, "Analyte_02", transform = "log2", levels = "Sample")
```

## PCA

`runPCA()` applies the missing-value filtering, imputation, transform,
centring and scaling, and returns the fit together with a readable trace of
what it did. `plotPCA()` draws the scores.

Blanks are dropped first. They are diagnostics rather than measurements, and
leaving them in gives them a say in the components: a blank sits far from every
real sample, so it can dominate the variance the PCA is trying to describe.
`runPCA()` does not filter them for you, so the subset is explicit.

```{r qc-pca}
syn_bio <- subsetDataset(syn, conditions = list(Sample_type = c("Sample", "QC")))

pca <- runPCA(syn_bio, impute = "gaussian", impute_frac = 0.2,
               transform = "log2", feat_na_max = 0.6,
               sample_na_max = 0.6)
pca$steps
```

The colouring is the whole point: the same scores answer a different question
depending on what you colour them by. The report draws one plot per column
listed under `qc_pca.color_by`.

```{r qc-pca-type, fig.height = 4.2}
plotPCA(pca, syn_bio, colour_by = "Sample_type")
```

Colouring by chromatographic batch is the sharpest of these -- if the batches
separate, the variance driving component 1 is analytical rather than
biological. This dataset had a batch shift planted in it, so here is the same
PCA before and after `correctBatch()`:

```{r qc-pca-batch, fig.height = 3.4}
pca_by_batch <- function(x, title){
  x <- subsetDataset(x, conditions = list(Sample_type = c("Sample", "QC")))
  p <- runPCA(x, impute = "gaussian", impute_frac = 0.2,
               transform = "log2")
  plotPCA(p, x, colour_by = "Chrom_Batch") +
    ggplot2::labs(title = title)
}

cowplot::plot_grid(
  pca_by_batch(de_uncorr, "Before batch correction"),
  pca_by_batch(de_bc,     "After batch correction"),
  nrow = 1)
```

## Loadings

The scores say *whether* samples separate; `plotLoadings()` says *which
compounds* drove it. Colouring by the enzymatic pathway is what turns a ranked
list into something interpretable -- the palette is shared with the heatmap
annotation, so a class is the same colour in every figure.

```{r qc-loadings, fig.width = 7, fig.height = 5}
plotLoadings(pca, syn_bio, colour_by = "Enzymatic_pathway")
```

## What was reported, and what was dropped

```{r qc-features}
dt(summariseFeatures(syn, out[[2]]), rownames = FALSE)
```

# Stage 4: statistics

## Selecting the biological samples

`subsetDataset()` slices by any `sample_meta` condition, which is how the
study samples are separated from the QCs and blanks before testing:

```{r stats-subset}
samples <- subsetDataset(syn,
                          conditions = list(Sample_type = "Sample"))

table(samples$sample_meta$Treatment, samples$sample_meta$Type)
```

## Group summaries

`summariseGroups()` is the table the box and bar plots are drawn from, and it
is written to the `summary` sheet of the stats workbook. Both SD and SE are
returned: SD describes the spread of the animals, SE how well the mean is
pinned down, and the choice belongs to the reader.

```{r stats-summary}
gs <- summariseGroups(samples, "Treatment")
dt(cbind(gs[, 1:3], round(gs[, 4:6], 3)), rownames = FALSE)
```

## Comparisons

`runStats()` runs the comparisons, correlations, linear models and metabolite
ratios described by the `stats_report:` block of the YAML and returns tidy
tables. The method is named explicitly per comparison: `welch` (the default),
`student` or `wilcoxon` where a factor has two levels, `anova` or `kruskal`
where it has three or more, the latter carrying Tukey and pairwise-Wilcoxon
post-hoc tests. Nothing is picked from the data itself, and p-values are
FDR-adjusted within each family.

Naming the test rather than inferring it is deliberate: targeted panels are
frequently run on small groups where normality cannot be taken on trust, so the
choice between a parametric and a non-parametric test is a real one. Making it
in the YAML, before the data is seen, keeps that judgement visible and stops a
more convenient result being selected after the fact.

```{r stats-run}
params <- list(
  sig_threshold = 0.05,
  comparisons = list(enabled = TRUE, entries = list(
    list(name    = "PBS_vs_HDM",
         method  = "welch",
         compare = list(factor = "Treatment",
                        levels = c("PBS", "HDM"))))),
  ion_ratios = list(enabled = TRUE, ratios = list(
    list(num = "Analyte_02", den = "Analyte_05",
         label = "A02/A05"))))

st <- runStats(samples, params)
```

Because the effects in this dataset were planted, the recovered fold changes
can be checked against the truth:

```{r stats-truth}
tt <- st$stats[st$stats$method == "welch",
               c("feature", "log2FC", "p_value", "p_adj")]
truth <- as.data.frame(syn$variable_meta)[, c("Compound",
                                              "Simulated_log2FC")]
chk <- merge(tt, truth, by.x = "feature", by.y = "Compound")
chk <- chk[order(-abs(chk$Simulated_log2FC), chk$p_value), ]

dt(cbind(chk[, 1], round(chk[, -1], 4)), rownames = FALSE, page = 8)
```

The five analytes with a non-zero `Simulated_log2FC` should carry the largest
recovered effects and the smallest p-values; the other fifteen should scatter
around zero.

## Per-compound plots

```{r stats-box, fig.width = 5, fig.height = 4}
plotBoxplot(samples, "Analyte_02", "Treatment",
             value_label = "ng/mL")
```

`type = "bar"` draws the group means and error bars from `summariseGroups()`,
so the figure and the `summary` sheet cannot drift apart:

```{r stats-bar, fig.width = 5, fig.height = 4}
plotBoxplot(samples, "Analyte_02", "Treatment", type = "bar",
             error_bar = "SE", value_label = "ng/mL")
```

## Volcano

Effect size against significance. A large fold change on three noisy animals is
not a finding, and a tiny but reproducible shift is rarely interesting either;
the compounds worth attention sit in the upper corners.

```{r stats-volcano, fig.width = 5.5, fig.height = 5}
plotVolcano(st, "PBS_vs_HDM", sig_threshold = 0.05)
```

## Heatmap

Where the comparisons test one compound at a time, the heatmap shows the whole
panel at once -- which is how you see that a treatment moved an entire pathway
rather than a few compounds that happened to clear a threshold.

```{r stats-heatmap, fig.width = 7.5, fig.height = 6}
plotHeatmap(samples,
             color_samples_by  = "Treatment",
             group_features_by = "Enzymatic_pathway",
             transform         = "log2",
             title             = "All analytes")
```

### When the sample axis runs out

`plotHeatmap()` never averages -- every sample is a column -- so use it for as
long as those columns stay legible, which in practice means two groups. Past
three, cells get narrow, the eye stops resolving individual samples, and the
usual retreat is a group-mean heatmap. That buys the space back by discarding
exactly the thing needed to trust it: whether the group agreed.

`plotGroupHeatmap()` is the same figure -- same annotation bars, same
palette, same blocks and gaps -- with one thing changed. The **hue** is the
group mean, so it reads as compactly as any group-mean heatmap; the
**opacity** of each sample's cell is that sample's contribution to it. A block
that is uniformly solid is an effect the whole group shares; a block carrying
one opaque cell among faded ones is a mean resting on a single sample.

```{r stats-group-heatmap, fig.width = 7.5, fig.height = 6}
plotGroupHeatmap(samples,
                   color_samples_by  = "Treatment",
                   group_features_by = "Enzymatic_pathway",
                   transform         = "log2",
                   title             = "All analytes, group means")
```

"Contribution" is measured against the mean of *all* samples, not against the
group's own members, so it is not the same as "has the largest value". If a
group holds one very high sample and three low ones, and the three sit further
from the overall centre than the one sits above it, then the group mean is
negative, the three produced it, and the high sample is the one opposing it.
The opacity answers "did this sample push its group where the group ended up",
not "is this sample large".

The two are shown together here to make the difference visible. **A report
renders one or the other, never both** -- `heatmap: style:` takes `samples`,
`group_means`, or `auto`, and `auto` (the default) picks `group_means` from
three groups upwards and `samples` at two.

## Feature correlations

Which compounds move *together*, which in a targeted panel usually means they
share a branch of the pathway -- so blocks along the diagonal are the pathway
structure showing itself.

```{r stats-corr, fig.width = 6.5, fig.height = 6}
cp <- list(correlations = list(enabled = TRUE, entries = list(
  list(name = "pairs", methods = "spearman",
       subsets = list(list())))))

plotCorrelation(runStats(samples, cp), samples,
                 group_by = "Enzymatic_pathway")
```

## Metabolite ratios

These are ratios between metabolites, used to infer enzymatic activity (not the
qualifier-to-quantifier transition ratios). The ratio of a product to its
precursor tracks flux through a pathway more directly than either compound
alone, because it cancels the shared variation in how much substrate was there.
`runStats()` returns the per-sample ratios with a group test attached, so they
reach the `ion_ratios` sheet as well as the report.

```{r stats-ratios}
ir <- st$ion_ratios[!is.na(st$ion_ratios$comparison), ]
dt(cbind(ir[, c("ratio", "comparison", "sample", "group")],
         round(ir[, c("value", "p_value")], 4)),
   rownames = FALSE)
```

# Driving it from a YAML config

In routine use none of the above is typed out. One YAML file describes the data
source, every processing step and both reports, and one call runs it:

```{r run-yaml, eval = FALSE}
res <- runMRManalyzeR("path/to/your_config.yml")
```

The bundled reference configuration is a commented template to copy and edit:

```{r config-path}
system.file("extdata", "example_config.yml", package = "MRManalyzeR")
```

Its `PeakMatrixProcessing:` block is a one-to-one map of stage 2 -- each key
switches a step on or names the metadata column it should use:

```yaml
PeakMatrixProcessing:
  execute: True                  # False = skip, reuse existing RDS

  name_col: Name                 # sample id (matches the matrix rows)
  include_col: Include           # sample include-filter column
  include_value: "YES"

  tl_data:                       # TargetLynx source (or skyline_data:)
    enabled: True
    data_tab_names: [lcms_data]
    tl_headers: ["ID", "Name", "Area", "ng/mL", "Response", "S/N"]
    snr: 3                       # S/N threshold | False

  blank_filter: 3                # n x blank mean | False
  blank_head: 'Sample_type'
  blank_name: 'Blank'

  filter_features:               # detection filter; omit or enabled: False
    enabled: True                #   to skip
    min_frac: 0.5
    method: within               # across | within | QC
    group_col: Treatment         # 'within' only
    sample_col: Sample_type      # QCs and blanks are excluded from the
    sample_labels: ["Sample"]     #   fraction but kept in the dataset

  filter_samples:                # runs after filter_features, on the
    enabled: True                #   reduced feature set
    max_na: 0.8
    sample_col: Sample_type      # only study samples are ever removed
    sample_labels: ["Sample"]

  batch_correction: True         # True | False | feature_metadata YES/NO column
  bc_qc_label: "QC"
  bc_factor_name: "Sample_type"
  bc_header: "Chrom_Batch"

  normalize: protein_ug          # False | sample_metadata column

  adjust_conc:
    enabled: True
    starting_vol_col: starting_vol_uL
    sample_vol_col:   sample_volume_uL
    cal_vol_col:      cal_vol_uL
    sample_IS_col:    sample_IS_vol_uL
    cal_IS_col:       cal_IS_vol_uL

  replace_MVs: 0.5               # False | fill ceiling (x feature min); last
  impute_method: gaussian        # gaussian (random draws) | constant
  impute_seed: 42                # same seed, same fills on every rerun
```

`data_quality_report:` and `stats_report:` configure stages 3 and 4 in the same
declarative way -- which column identifies QCs, which factors to compare, which
plots to draw. See the bundled file for the fully commented set.

Outputs are written next to your data: the processed matrix as `.xlsx` and
`.RDS` (imputed when `replace_MVs` is set, with the measured values as a
`matrix_measured` sheet and a `_measured.RDS`), a `_stats.xlsx` carrying
`stats`, `correlations`, `linear_models`, `ion_ratios` and `summary`, the two
self-contained HTML reports, and a text file recording the exact parameters
used.

The stats workbook opens on a `key` sheet defining every column heading it
contains -- what `adjust_family` covers, how `log2FC` is signed, which effect
size `effect_type` names.

## Merging several acquisition panels

Where a study is split across methods, for example a polar-acid and a
polar-basic panel, each is processed on its own with `runMRManalyzeR()` first,
and the resulting `.RDS` files are merged:

```{r run-combine, eval = FALSE}
file.copy(
  system.file("extdata", "example_combine_config.yml",
              package = "MRManalyzeR"),
  "combine_config.yml")

res <- runMRManalyzeRCombine("combine_config.yml")
```

Two things decide whether a merge succeeds: `sample_id_col` must name an
identifier that is stable across the panels (not the acquisition `Name`, which
usually carries a method-specific prefix), and each dataset's `tag` prefixes
its feature names so that a compound measured on two methods does not collide.

# References {#references}

Broadhurst D. *et al.* (2018). Guidelines and considerations for the use of
system suitability and quality control samples in mass spectrometry assays
applied in untargeted clinical metabolomic studies. *Metabolomics*.
doi:[10.1007/s11306-018-1367-3](https://doi.org/10.1007/s11306-018-1367-3)
-- source of the dispersion-ratio reference point used in the QC section.

Kolmert J. *et al.* (2018). Prominent release of lipoxygenase generated
mediators in a murine house dust mite-induced asthma model.
*Prostaglandins & Other Lipid Mediators*.
doi:[10.1016/j.prostaglandins.2018.05.005](https://doi.org/10.1016/j.prostaglandins.2018.05.005)
-- source of the bundled example data.

Lloyd G.R., Jankevics A., Weber R.J.M. (2020). struct: an R/Bioconductor-based
framework for standardized metabolomics data analysis. *Bioinformatics*.
doi:[10.1093/bioinformatics/btaa1031](https://doi.org/10.1093/bioinformatics/btaa1031)
-- the `DatasetExperiment` container used throughout.

Wilkinson M.D. *et al.* (2016). The FAIR Guiding Principles for scientific data
management and stewardship. *Scientific Data* **3**, 160018.
doi:[10.1038/sdata.2016.18](https://doi.org/10.1038/sdata.2016.18)
-- the reproducibility principles the configuration-driven workflow follows.

# Session info

```{r session-info}
sessionInfo()
```
