---
title: "MRManalyzeR — Data Quality Report"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    code_folding: hide
    css: !expr system.file("rmd","IMP_ReportCSS.css", package="MRManalyzeR")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = FALSE,
  warning     = FALSE,
  message     = FALSE,
  fig.align   = "center",
  fig.retina  = 3
)

library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r unpack-params}
# `report_pars` is injected by runMRManalyzeR().
# Falls back to legacy uva_pars / mva_pars if running outside that wrapper.
if(!exists("report_pars") || is.null(report_pars))
  report_pars = if(exists("uva_pars")) uva_pars else mva_pars

`%||%` = function(a, b) if(is.null(a)) b else a

# Colour scheme (shared by every plot). Set once here rather than passed to
# each plotting function, so a report is internally consistent by default.
.cols = report_pars$colours %||% list()
options(MRManalyzeR.palette   = .cols$qualitative %||% "hat",
        MRManalyzeR.diverging = .cols$diverging   %||% "heatmap2")

# -- Sample identification --
# One column classifies every row; qc_label / sample_labels name the values
# marking QC injections and biological samples. Every section references
# these, so the classifier column is set once, not per-section.
sample_type_head = report_pars$sample_type_head %||% "Sample_type"
qc_label         = report_pars$qc_label         %||% "QC"
sample_labels    = report_pars$sample_labels    %||% "Sample"

# Which sample types appear in the intensity-drift / Q-Q plots (default QC + samples)
intensity_plot_levels = report_pars$intensity_plot_levels %||%
                          unique(c(qc_label, sample_labels))

yaxis_quant  = report_pars$yaxis_quant %||% 0.95
yaxis_scalar = as.numeric(report_pars$yaxis_scalar %||% 1.3)

# sample_metadata column names -- configurable since studies name these
# differently; defaults match the common convention.
injection_order_head = report_pars$injection_order_head %||% "Injection_order"
sample_id_head        = report_pars$sample_id_head       %||% "Sample_ID"

units      = pmp_params$units %||% pmp_params$datatype %||%
             project_paths$datatype %||% ""
# Y-value label for point tooltips: the datatype units (ng/mL, Response, …),
# or "value" when unset. Coerced to a scalar (datatype may arrive as a list).
value_label = local({
  u = as.character(units)[1]
  if(!is.na(u) && nzchar(u)) u else "value"
})
bc_header  = pmp_params$bc_header

# Performance / layout knobs
plot_intensity_per_compound = isTRUE(report_pars$plot_intensity_per_compound %||% TRUE)
plot_qq_per_compound        = isTRUE(report_pars$plot_qq_per_compound        %||% TRUE)
# qq_transform may be a single value or a list — list form creates one tab per transform.
qq_transforms               = report_pars$qq_transform %||% c("none", "log2")
if(length(qq_transforms) == 1) qq_transforms = as.character(qq_transforms)
fig_scale                   = as.numeric(report_pars$fig_scale %||% 1)

# QC PCA config
qc_pca_include = isTRUE(report_pars$qc_pca$enabled  %||% FALSE)
qc_pca_color   = report_pars$qc_pca$color_by        %||% character(0)
# Default INCLUDES the QC sample type so the QC PCA shows QCs alongside
# samples (drift / batch / clustering checks). Override qc_pca.levels: in YAML.
qc_pca_levels  = report_pars$qc_pca$levels %||% unique(c(sample_labels, qc_label))
qc_pca_head    = report_pars$qc_pca$head   %||% sample_type_head
qc_pca_label   = report_pars$qc_pca$label           %||% "none"
qc_pca_label_factor = report_pars$qc_pca$label_factor %||% sample_id_head
qc_pca_ellipse = report_pars$qc_pca$ellipse         %||% "none"

# Preprocessing for QC PCA (with sensible defaults)
qc_pca_pp = report_pars$qc_pca$preprocess %||% list()
qc_pca_impute        = qc_pca_pp$impute        %||% "gaussian"
qc_pca_impute_frac   = as.numeric(qc_pca_pp$impute_frac %||% 0.5)
qc_pca_seed          = if("seed" %in% names(qc_pca_pp)) qc_pca_pp$seed else 42
qc_pca_transform     = qc_pca_pp$transform     %||% "log2"
qc_pca_center        = isTRUE(qc_pca_pp$center %||% TRUE)
qc_pca_scale         = isTRUE(qc_pca_pp$scale  %||% TRUE)
qc_pca_feat_na_max   = as.numeric(qc_pca_pp$feat_na_max   %||% 1)
qc_pca_sample_na_max = as.numeric(qc_pca_pp$sample_na_max %||% 1)

# Interactive plotting (plotly hover tooltips). Default ON. Set
# `data_quality_report.interactive_plots: False` in YAML to fall back to png.
interactive_plots = isTRUE(report_pars$interactive_plots %||% TRUE) &&
                    requireNamespace("plotly", quietly = TRUE)

.show_plot = function(p, lock_aspect = FALSE, legend_below = FALSE,
                      max_width = NULL, pixel_height = NULL){
  if(interactive_plots){
    w = plotly::ggplotly(p, tooltip = "text", height = pixel_height)
    if(isTRUE(lock_aspect))
      w = plotly::layout(w, yaxis = list(scaleanchor = "x", scaleratio = 1))
    if(isTRUE(legend_below)){
      w = plotly::layout(w,
                         margin = list(b = 170, t = 40),
                         legend = list(orientation = "h",
                                       x = 0.5, xanchor = "center",
                                       y = -0.40, yanchor = "top"))
      # Continuous colorbars (Injection_order, numeric Chrom_Batch) live on
      # each trace's `marker.colorbar`. Walk the figure data and only modify
      # traces that already carry a colorbar — discrete-color traces don't,
      # so they remain untouched (avoids the phantom-colorbar bug).
      for(.i in seq_along(w$x$data)){
        .mk = w$x$data[[.i]]$marker
        if(isTRUE(.mk$showscale) || !is.null(.mk$colorscale)){
          if(is.null(w$x$data[[.i]]$marker$colorbar))
            w$x$data[[.i]]$marker$colorbar = list()
          w$x$data[[.i]]$marker$colorbar$orientation = "h"
          w$x$data[[.i]]$marker$colorbar$x          = 0.5
          w$x$data[[.i]]$marker$colorbar$xanchor    = "center"
          w$x$data[[.i]]$marker$colorbar$y          = -0.30
          w$x$data[[.i]]$marker$colorbar$yanchor    = "top"
          w$x$data[[.i]]$marker$colorbar$len        = 0.7
          w$x$data[[.i]]$marker$colorbar$thickness  = 14
        }
      }
    }
    w = plotly::config(w, displaylogo = FALSE,
                       modeBarButtonsToRemove = c("lasso2d", "select2d"))
    tags = htmltools::as.tags(w, standalone = FALSE)
    deps = htmltools::findDependencies(tags)
    if(length(deps)) knitr::knit_meta_add(list(deps))
    html_str = as.character(tags)
    if(!is.null(max_width))
      html_str = sprintf('<div style="max-width:%s;margin:0 auto;">%s</div>',
                         max_width, html_str)
    cat(html_str)
  } else {
    print(p)
  }
}

```

# 1. Aim

Inspect data quality for acquired compounds. Check QCs, drift over injection
order, and confounding factors (batch, sample type) via PCA.

`r .n_imp = .n_imputed_total(combined_datamatrices); .is_imp = if(exists("values_imputed")) isTRUE(values_imputed) else .n_imp > 0; if(.is_imp) sprintf("Values: an **imputed** matrix - %s%s.", if(.n_imp > 0) sprintf("%d value(s) were filled in", .n_imp) else "how many values were filled in was not recorded", if(isTRUE(report_pars$use_imputed)) " (\x60use_imputed: True\x60)" else ", because the stored dataset predates the measured copy") else "Values: as **measured**, gaps included - nothing in this report is imputed except inside the QC PCA."`

# 2. Data import and summary

```{r import}
importData      = combined_datamatrices
metadata        = importData$sample_meta
data_matrix     = importData$data
compoundInfo    = importData$variable_meta
compound_names  = compoundInfo$Compound

DT::datatable(
  metadata %>%
    dplyr::group_by(.data[[sample_type_head]]) %>%
    dplyr::summarize(`Number of samples` = dplyr::n(), .groups = "drop")
)
```

# 3. Quality CVs

```{r cvs}
# Per-feature CV in QC vs samples (when QC samples present)
# CV_QC over QC rows; CV_sample over biological-sample rows only (NOT
# "everything non-QC", which would fold in blanks / calibration curves).
# runMRManalyzeR() computes the CVs on the measured values and stores them;
# use those, so the table matches the workbook whichever matrix this report
# was given. Recompute only for a dataset that has none.
qc_idx = which(metadata[[sample_type_head]] %in% qc_label)

.cv_cols = c("CV_QC", "CV_sample", "CV_sample_vs_QC")
.cv_de = if(all(.cv_cols %in% colnames(as.data.frame(combined_datamatrices$variable_meta))))
  combined_datamatrices else
  addCVMetrics(combined_datamatrices,
               sample_type_head = sample_type_head,
               qc_label         = qc_label,
               sample_labels    = sample_labels)

cv_table = as.data.frame(.cv_de$variable_meta)[
  , c("Compound", "CV_QC", "CV_sample", "CV_sample_vs_QC"), drop = FALSE]

if(length(qc_idx) < 2){
  cat(sprintf(
    "_No QC samples detected (metadata$%s in {%s} had %d row(s)). CV_QC column shown as NA._\n\n",
    sample_type_head, paste(qc_label, collapse = ", "), length(qc_idx)))
}

DT::datatable(cv_table, filter = "top",
              options = list(pageLength = 25, scrollX = TRUE))
```

# 4. Per-compound intensity vs injection order

`r if(!plot_intensity_per_compound) "_Skipped — set_ \x60plot_intensity_per_compound: True\x60 _in YAML to enable._"`

```{r intensity-dropdown, results="asis", eval=plot_intensity_per_compound, fig.width=10*fig_scale, fig.height=5*fig_scale}
make_safe_id = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
safe_ids = vapply(compound_names, make_safe_id, character(1))
# Alphabetise dropdown options for easier scanning (T3.5-7); plot blocks
# below still iterate in import order.
.alpha_idx        = order(compound_names)
.compound_alpha   = compound_names[.alpha_idx]
.safe_ids_alpha   = safe_ids[.alpha_idx]

cat('
<label for="compound_select" style="font-weight:600;">Select compound:</label>
<select id="compound_select" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showCompoundPlot(this.value)">
')
for(i in seq_along(.compound_alpha)){
  sel = if(i == 1) " selected" else ""
  cat(sprintf('<option value="%s"%s>%s</option>\n',
              .safe_ids_alpha[i], sel, .compound_alpha[i]))
}
cat('</select>
<script>
function showCompoundPlot(id){
  var plots = document.querySelectorAll(".compound_plot_block");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("plot_" + id);
  if (el) {
    el.style.display = "block";
    if (window.Plotly){
      var pls = el.querySelectorAll(".plotly-graph-div");
      for (var j=0; j<pls.length; j++) Plotly.Plots.resize(pls[j]);
    }
  }
}
</script>
')

for(i in seq_along(compound_names)){
  compound = compound_names[i]
  cid      = safe_ids[i]

  p = plotDrift(combined_datamatrices, compound,
                 sample_type_head     = sample_type_head,
                 levels               = intensity_plot_levels,
                 injection_order_head = injection_order_head,
                 sample_id_head       = sample_id_head,
                 value_label          = units,
                 yaxis_quant          = yaxis_quant,
                 yaxis_scalar         = yaxis_scalar)

  # Show the alphabetically-first compound by default (matches dropdown selection)
  display = if(safe_ids[i] == .safe_ids_alpha[1]) "block" else "none"
  cat(sprintf('<div class="compound_plot_block" id="plot_%s" style="display:%s;">\n',
              cid, display))
  .show_plot(p, legend_below = TRUE)
  cat("</div>\n")
}
```

# 5. Q-Q plot per compound (normality check) {.tabset}

`r if(!plot_qq_per_compound) "_Skipped — set_ \x60plot_qq_per_compound: True\x60 _in YAML to enable._"`

`r if(plot_qq_per_compound) sprintf("Sample-level normality assessment per compound. Filter applied to drop blanks/QCs/calibration: \x60%s\x60 _∈ {%s}_ → _%d samples kept (of %d in metadata). One tab per transform listed in_ \x60qq_transform\x60_:_ %s_. Caption shows Shapiro-Wilk W and p (3 ≤ n ≤ 5000)._", sample_type_head, paste(intensity_plot_levels, collapse=", "), sum(metadata[[sample_type_head]] %in% intensity_plot_levels), nrow(metadata), paste(qq_transforms, collapse=", "))`

```{r qq-dropdown, results="asis", eval=plot_qq_per_compound, fig.width=6*fig_scale, fig.height=5*fig_scale}
make_safe_id_qq = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
# Iterate compounds in import order for the plot blocks; the dropdown lists
# them alphabetically (T3.5-7) for easier scanning.
safe_ids_qq      = vapply(compound_names, make_safe_id_qq, character(1))
alpha_idx        = order(compound_names)
compound_alpha   = compound_names[alpha_idx]
safe_ids_alpha   = safe_ids_qq[alpha_idx]

for(tr in qq_transforms){
  tr_safe = make_safe_id_qq(tr)
  cat(sprintf("\n## %s {-}\n\n", tr))

  cat(sprintf('
<label for="qq_select_%s" style="font-weight:600;">Select compound:</label>
<select id="qq_select_%s" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showQQPlot_%s(this.value)">
', tr_safe, tr_safe, tr_safe))
  for(i in seq_along(compound_alpha)){
    sel = if(i == 1) " selected" else ""
    cat(sprintf('<option value="%s"%s>%s</option>\n',
                safe_ids_alpha[i], sel, compound_alpha[i]))
  }
  cat(sprintf('</select>
<script>
function showQQPlot_%s(id){
  var plots = document.querySelectorAll(".qq_plot_block_%s");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("qq_%s_" + id);
  if (el) el.style.display = "block";
}
</script>
', tr_safe, tr_safe, tr_safe))

  for(i in seq_along(compound_names)){
    compound = compound_names[i]
    cid      = safe_ids_qq[i]

    p = plotQQ(combined_datamatrices, compound, transform = tr,
                sample_type_head = sample_type_head,
                levels           = intensity_plot_levels,
                value_label      = units)

    # Each compound's plot div uses the dropdown safe_id (alpha-list value);
    # its leading style remembers which compound is the alpha-first one.
    display = if(safe_ids_qq[i] == safe_ids_alpha[1]) "block" else "none"
    cat(sprintf('<div class="qq_plot_block_%s" id="qq_%s_%s" style="display:%s;">\n',
                tr_safe, tr_safe, cid, display))
    print(p)
    cat("</div>\n")
  }
}
```

# 6. QC PCA (drift / batch / confounder inspection)

`r if(!qc_pca_include) "_Skipped — set_ \x60qc_pca.enabled: True\x60 _in YAML to enable._"`

`r if(qc_pca_include) sprintf("PCA computed once on the filtered subset; one plot per coloring listed in_ \x60qc_pca.color_by\x60_:_ %s.", paste(qc_pca_color, collapse=", "))`

```{r qc-pca, results="asis", eval=qc_pca_include && length(qc_pca_color) > 0, fig.width=7*fig_scale, fig.height=6*fig_scale}
# Subset to requested sample types
keep = metadata[[qc_pca_head]] %in% qc_pca_levels
X = as.matrix(data_matrix[keep, , drop = FALSE])
# Values from the exported matrix: what replace_MVs imputed, so runPCA()
# fills only what is still missing.
.P = get0("pca_datamatrices")
if(!is.null(.P)){
  .P = as.matrix(as.data.frame(.P$data))
  if(all(rownames(X) %in% rownames(.P)) && all(colnames(X) %in% colnames(.P)))
    X = .P[rownames(X), colnames(X), drop = FALSE]
}
smeta_pca = metadata[keep, , drop = FALSE]

# Ensure shared rownames on X and smeta_pca so we can re-align after the
# PCA pipeline drops samples via `sample_na_max`.
.qc_rn = if(!is.null(rownames(X)) && !anyDuplicated(rownames(X))) {
  rownames(X)
} else {
  paste0("qc_pca_", seq_len(nrow(X)))
}
rownames(X)         = .qc_rn
rownames(smeta_pca) = .qc_rn

res_pca = runPCA(X,
                           impute        = qc_pca_impute,
                           impute_frac   = qc_pca_impute_frac,
                           seed          = qc_pca_seed,
                           transform     = qc_pca_transform,
                           center        = qc_pca_center,
                           scale         = qc_pca_scale,
                           feat_na_max   = qc_pca_feat_na_max,
                           sample_na_max = qc_pca_sample_na_max)

# Realign metadata to the samples that survived sample_na_max filtering.
if(!is.null(res_pca$pr))
  smeta_pca = smeta_pca[rownames(res_pca$pr$x), , drop = FALSE]

# Show what was actually applied
cat("**Preprocessing pipeline applied:**\n\n")
for(s in res_pca$steps) cat(sprintf("- %s\n", s))
cat(sprintf("\n_Final matrix: %d samples × %d features._\n\n",
            res_pca$n_samples, res_pca$n_features))

if(is.null(res_pca$pr)){
  cat(sprintf("\n_PCA could not be fit on cleaned matrix (n_samples=%d, n_features=%d)._\n",
              res_pca$n_samples, res_pca$n_features))
} else {
  # One tab per coloring listed in qc_pca.color_by
  cat("\n## QC PCA scatter plots {.tabset .tabset-fade}\n\n")

  for(col_name in qc_pca_color){
    cat(sprintf("\n### Coloured by %s {-}\n\n", col_name))

    if(!col_name %in% colnames(smeta_pca)){
      cat(sprintf("_Column `%s` not in sample_meta — skipped._\n\n", col_name))
      next
    }

    # Injection order is stored as character in some workbooks but is really a
    # sequence, so force a continuous scale for it; everything else is decided
    # from the column's own type.
    p = plotPCA(res_pca, smeta_pca, colour_by = col_name,
                 label         = qc_pca_label,
                 label_factor  = qc_pca_label_factor,
                 ellipse       = qc_pca_ellipse,
                 numeric       = if(col_name == injection_order_head) TRUE
                                 else NA)

    .show_plot(p, lock_aspect = TRUE, legend_below = TRUE,
               max_width = "650px", pixel_height = 650)
    cat("\n\n")
  }
}
```

# 7. Final considerations

Every compound the run started with, and whether it survived into the reported
matrix. Exclusions include S/N or LOD/LOQ failures that left a feature entirely
missing, features whose `Processing_name` had no rows in the source data, and
source columns with no `feature_metadata` entry — filter the `Status` column to
see one group at a time.

```{r feature-summary}
feature_summary = summariseFeatures(combined_datamatrices, removed_features)

DT::datatable(feature_summary, filter = "top",
              options = list(pageLength = 25, scrollX = TRUE))
```

```{r feature-summary-note, results="asis"}
if(is.null(removed_features) || nrow(removed_features) == 0)
  cat("_No exclusions recorded (typical when PeakMatrixProcessing.execute",
      "was False, since nothing was reprocessed)._")
```

# 8. Session info

```{r session}
tryCatch(sessionInfo(), error = function(e){
  cat("sessionInfo() failed (likely a malformed DESCRIPTION in an installed package):\n")
  cat(conditionMessage(e), "\n\n")
  cat("R version:", R.version$version.string, "\n")
  cat("Platform :", R.version$platform, "\n")
  cat("Loaded packages:\n")
  cat(paste(" ", sort(loadedNamespaces()), collapse = "\n"), "\n")
})
```
