---
title: "MRManalyzeR — Statistics Report"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    code_folding: hide
    css: !expr system.file("rmd","IMP_ReportCSS.css", package="MRManalyzeR")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = FALSE,
  warning     = FALSE,
  message     = FALSE,
  fig.align   = "center",
  fig.retina  = 3
)

library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r unpack-params}
if(!exists("report_pars") || is.null(report_pars))
  report_pars = if(exists("mva_pars")) mva_pars else uva_pars

`%||%` = function(a, b) if(is.null(a)) b else a

# Colour scheme (shared by every plot). Set once here rather than passed to
# each plotting function, so a report is internally consistent by default.
.cols = report_pars$colours %||% list()
options(MRManalyzeR.palette   = .cols$qualitative %||% "hat",
        MRManalyzeR.diverging = .cols$diverging   %||% "heatmap2")

# -- Sample identification (shared across all sections) --
# One column classifies every row; sample_labels names the value(s) marking
# biological samples (vs blanks / QCs / calibration standards).
# sample_id_head is the per-sample identifier used for hover tooltips and
# heatmap row labels (and the default for pca.label_factor).
sample_type_head = report_pars$sample_type_head %||% "Sample_type"
sample_labels    = report_pars$sample_labels    %||% "Sample"
sample_id_head   = report_pars$sample_id_head   %||% "Sample_ID"

# Stats blocks
# Each section is a block with a section-level `enabled:` toggle and an
# `entries:` list; .section_entries() honours the toggle and returns the
# entries (or list() when disabled/absent).
comparisons   = .section_entries(report_pars$comparisons,   "comparisons")
correlations  = .section_entries(report_pars$correlations,  "correlations")
linear_models = .section_entries(report_pars$linear_models, "linear_models")

p_adjust      = report_pars$p_adjust      %||% "BH"
sig_threshold = report_pars$sig_threshold %||% 0.05

# Plot config
fig_scale       = as.numeric(report_pars$fig_scale %||% 1)
boxplot_faceted = isTRUE(report_pars$boxplot$faceted %||% TRUE)
pca_include     = isTRUE(report_pars$pca$enabled   %||% TRUE)
pca_label       = report_pars$pca$label            %||% "none"
pca_label_factor = report_pars$pca$label_factor    %||% sample_id_head
pca_ellipse     = report_pars$pca$ellipse          %||% "group"
pca_ellipse_conf = as.numeric(report_pars$pca$ellipse_confidence %||% 0.95)

# PCA preprocessing (with sensible defaults)
pca_pp = report_pars$pca$preprocess %||% list()
pca_impute        = pca_pp$impute      %||% "gaussian"
pca_impute_frac   = as.numeric(pca_pp$impute_frac %||% 0.5)
pca_seed          = if("seed" %in% names(pca_pp)) pca_pp$seed else 42
pca_transform     = pca_pp$transform   %||% "log2"
pca_center        = isTRUE(pca_pp$center %||% TRUE)
pca_scale         = isTRUE(pca_pp$scale  %||% TRUE)
pca_feat_na_max   = as.numeric(pca_pp$feat_na_max   %||% 1)
pca_sample_na_max = as.numeric(pca_pp$sample_na_max %||% 1)

# Component pair to plot (default PC1 vs PC2). Length-2 vector of integers.
pca_components    = report_pars$pca$components %||% c(1, 2)
pca_pc_x          = as.integer(pca_components[1])
pca_pc_y          = as.integer(pca_components[length(pca_components)])

# Optional loadings plot below scores; coloured by a feature_metadata column.
pca_loadings_show = isTRUE(report_pars$pca$loadings$enabled %||% TRUE)
pca_loadings_color_by = report_pars$pca$loadings$color_by %||% NULL

# Optional extra subset for the GLOBAL PCA tab (multi-key, AND'd together).
# This is on top of the biological_filter, e.g. pca.subset: { Lobe: Left }
# to inspect only the left-lobe samples in the global PCA.
pca_global_subset = report_pars$pca$subset %||% NULL

volcano_include   = isTRUE(report_pars$volcano$enabled %||% TRUE)
volcano_top_n     = as.integer(report_pars$volcano$top_n_label %||% 10)
# Volcano y-axis: use raw or BH-adjusted p (default = unadjusted, per request)
volcano_use_adj   = isTRUE(report_pars$volcano$use_adjusted %||% FALSE)

# Heatmap (z-score, samples x features). Shared render settings live at the
# heatmap: root and are INHERITED by the general / top_changing / split_by_class
# sub-blocks; any can be OVERRIDDEN per block (resolved via .resolve_hm below).
heatmap_include            = isTRUE(report_pars$heatmap$enabled %||% TRUE)
heatmap_color_samples_by   = report_pars$heatmap$color_samples_by  %||%
                             report_pars$global_summary_factor %||% sample_type_head
heatmap_group_features_by  = report_pars$heatmap$group_features_by %||% NULL
heatmap_transform          = report_pars$heatmap$transform %||% "log2"
heatmap_scale              = report_pars$heatmap$scale     %||% "z"
heatmap_cluster_rows       = isTRUE(report_pars$heatmap$cluster_rows %||% FALSE)
heatmap_cluster_within     = isTRUE(report_pars$heatmap$cluster_within_groups %||% TRUE)
# Orientation: which axis carries samples. "samples_y" (default) = samples on
# rows / y-axis, features on columns / x-axis. "samples_x" transposes. "auto"
# puts the larger dimension on the y-axis.
heatmap_orientation        = tolower(report_pars$heatmap$orientation %||% "samples_y")
# Feature-label density: show | small | hide. "small" shrinks the feature-axis
# labels so a large panel stays readable via the class colour annotation bar;
# "hide" drops them entirely.
heatmap_feature_labels     = tolower(report_pars$heatmap$feature_labels %||% "small")
heatmap_style              = tolower(report_pars$heatmap$style %||% "auto")
heatmap_alpha_floor        = as.numeric(report_pars$heatmap$alpha_floor %||% 0.6)
# Cell edge in points. Absolute units keep cells square inside the canvas the
# report stretches the heatmap grob into; proportional cells would not be.
heatmap_cell_size = report_pars$heatmap$cell_size %||% "auto"
# How far from square a cell may get when cell_size is "auto". 1 forces exact
# squares; 1.5 lets a very lopsided panel use the canvas without the cells
# looking stretched.
heatmap_cell_aspect = as.numeric(report_pars$heatmap$cell_aspect %||% 1.5)

# Root default for the 4.6 comparison-tab ranking metric (top_changing may override).
heatmap_rank_by            = tolower(report_pars$heatmap$rank_by %||% "p_adj")

# Validators: coerce out-of-range values back to the safe default.
.chk_scope  = function(x) if(x %in% c("global","comparisons","all")) x else "all"
.chk_orient = function(x) if(x %in% c("samples_y","samples_x","auto")) x else "samples_y"
.chk_style  = function(x) if(x %in% c("auto","samples","group_means")) x else "auto"
# 'auto' picks per heatmap: with two groups every sample is legible side by
# side, so plotHeatmap() shows strictly more; from three the cells narrow and
# the eye stops resolving individual samples, which is where the group-mean
# rendering with per-sample opacity is worth the unfamiliar encoding. Only ONE
# style is ever drawn for a given heatmap - the two are alternatives, not a
# before-and-after pair.
# Braces are load-bearing: an unbraced `if(...) x` followed by `else` on the
# next line parses as a complete `if` plus a stray `else`, which is a syntax
# error at chunk level and takes the whole report down with it.
.pick_style = function(style, n_groups){
  if(!identical(style, "auto")) return(style)
  if(n_groups >= 3) "group_means" else "samples"
}
.chk_labels = function(x) if(x %in% c("show","small","hide")) x else "small"
heatmap_orientation    = .chk_orient(heatmap_orientation)
heatmap_feature_labels = .chk_labels(heatmap_feature_labels)
heatmap_style          = .chk_style(heatmap_style)

# Resolve a sub-block's render settings against the root defaults — the full
# settings list consumed by .draw_heatmap(). Every setting is overridable per
# block (block$x %||% root_x).
.resolve_hm = function(block){
  block = block %||% list()
  list(
    color_samples_by  = block$color_samples_by  %||% heatmap_color_samples_by,
    group_features_by = block$group_features_by %||% heatmap_group_features_by,
    transform         = tolower(block$transform %||% heatmap_transform),
    scale             = tolower(block$scale     %||% heatmap_scale),
    cluster_rows      = isTRUE(block$cluster_rows          %||% heatmap_cluster_rows),
    cluster_within    = isTRUE(block$cluster_within_groups %||% heatmap_cluster_within),
    orientation       = .chk_orient(tolower(block$orientation    %||% heatmap_orientation)),
    feature_labels    = .chk_labels(tolower(block$feature_labels %||% heatmap_feature_labels)),
    cell_size         = block$cell_size %||% heatmap_cell_size,
    cell_aspect       = as.numeric(block$cell_aspect %||% heatmap_cell_aspect),
    style             = .chk_style(tolower(block$style %||% heatmap_style)),
    alpha_floor       = as.numeric(block$alpha_floor %||% heatmap_alpha_floor)
  )
}

# 4.5 general — all features retained.
heatmap_general       = report_pars$heatmap$general %||% list()
heatmap_general_on    = isTRUE(heatmap_general$enabled %||% TRUE)
heatmap_general_scope = .chk_scope(tolower(heatmap_general$scope %||% "all"))
heatmap_general_cfg   = .resolve_hm(heatmap_general)

# 4.6 top-changing — top-N changing features per class. Comparison tabs rank by
# `rank_by` (block %||% root); the Global tab ranks by separation across
# color_samples_by (no single 2-group contrast exists globally).
heatmap_topchg        = report_pars$heatmap$top_changing %||% list()
heatmap_topchg_on     = isTRUE(heatmap_topchg$enabled %||% FALSE)
heatmap_topchg_scope  = .chk_scope(tolower(heatmap_topchg$scope %||% "comparisons"))
heatmap_topchg_n      = as.integer(heatmap_topchg$n_per_class %||% 10)
heatmap_topchg_rank   = tolower(heatmap_topchg$rank_by %||% heatmap_rank_by)
heatmap_topchg_cfg    = .resolve_hm(heatmap_topchg)

# 4.7 by class — one heatmap per subclass_col value, or one COMBINED heatmap
# across the selected values (multi_class). `subclass_values` focuses specific
# values (empty/NULL = all).
heatmap_split         = report_pars$heatmap$split_by_class %||% list()
heatmap_split_on      = isTRUE(heatmap_split$enabled %||% FALSE)
heatmap_split_scope   = .chk_scope(tolower(heatmap_split$scope %||% "global"))
heatmap_subclass_col  = heatmap_split$subclass_col %||% NULL
heatmap_subclass_vals = heatmap_split$subclass_values %||% NULL
heatmap_split_multi   = tolower(heatmap_split$multi_class %||% "combined")
if(!heatmap_split_multi %in% c("combined","tabs")) heatmap_split_multi = "combined"
heatmap_split_cfg     = .resolve_hm(heatmap_split)

# Biological-sample filter (drops blanks/QCs from global summary + global PCA).
# Defaults to sample_type_head ∈ sample_labels. Set biological_filter.head: ~ to disable.
bio_filter        = report_pars$biological_filter %||%
                    list(head = sample_type_head, levels = sample_labels)
bio_filter_head   = bio_filter$head
bio_filter_levels = bio_filter$levels

# General per-compound boxplot/barplot (independent of comparisons)
boxplot_enabled = isTRUE(report_pars$boxplot$enabled %||% FALSE)
boxplot_factors = report_pars$boxplot$factors %||%
                   c(report_pars$boxplot_factor1, report_pars$boxplot_factor2)
boxplot_factors = boxplot_factors[!is.null(boxplot_factors) & nzchar(boxplot_factors)]
boxplot_display = report_pars$boxplot$display %||% "tabs"      # combined | tabs
boxplot_levels  = report_pars$boxplot$levels %||% sample_labels
boxplot_head    = report_pars$boxplot$head   %||% sample_type_head

# Plot type — applies to BOTH section 4.2 (faceted across comparisons) and 4.3
# (general by-factor). Values: "box" | "bar" | "both". When "both", boxplot
# renders first then barplot underneath. Error bars on barplots controlled by
# `boxplot.error_bar`: "SD" | "SE" | "none".
plot_type      = tolower(report_pars$boxplot$plot_type %||%
                          report_pars$plot_type %||% "box")
if(!plot_type %in% c("box", "bar", "both")) plot_type = "box"
plot_error_bar = toupper(report_pars$boxplot$error_bar %||%
                          report_pars$error_bar %||% "SD")
if(!plot_error_bar %in% c("SD", "SE", "NONE")) plot_error_bar = "SD"

# Per-comparison boxplot toggle (faceted across comparisons)
plot_boxplot_per_comparison = isTRUE(report_pars$plot_boxplot_per_compound %||% TRUE)
yaxis_quant    = report_pars$yaxis_quant    %||% 0.95
yaxis_scalar = as.numeric(report_pars$yaxis_scalar %||% 1.3)

# Sample-level summary bar plots; default group/colour by the sample-type classifier.
global_summary_factor = report_pars$global_summary_factor %||% sample_type_head

units = pmp_params$units %||% pmp_params$datatype %||% project_paths$datatype %||% ""

# Y-value label for point tooltips: the datatype units (ng/mL, Response, …),
# or "value" when unset. Coerced to a scalar (datatype may arrive as a list).
value_label = local({
  u = as.character(units)[1]
  if(!is.na(u) && nzchar(u)) u else "value"
})

# Ion ratios (enzyme activity surrogates etc.)
ion_ratios_list    = ((report_pars$ion_ratios %||% list())$ratios) %||% list()
ion_ratios_enabled = length(ion_ratios_list) > 0 &&
                     isTRUE((report_pars$ion_ratios %||% list())$enabled %||% TRUE)

# Interactive plotting (plotly hover tooltips). Default ON. Set
# `stats_report.interactive_plots: False` in YAML to fall back to static png.
interactive_plots = isTRUE(report_pars$interactive_plots %||% TRUE) &&
                    requireNamespace("plotly", quietly = TRUE)

# Render a ggplot — interactive (plotly with `text` tooltip) when
# interactive_plots is TRUE, otherwise standard static print. Use in place of
# print(p) inside results="asis" chunks.
#
# Why the explicit dependency dance: in a `results="asis"` chunk that
# interleaves cat()'d HTML with widgets, `print(htmltools::tagList(w))` does
# NOT reliably register the plotly.js / htmlwidgets.js dependencies into the
# document <head>, so widgets render blank. Building tags explicitly and
# calling knit_meta_add() forces knitr to inject the deps once per render.
.show_plot = function(p, lock_aspect = FALSE, legend_below = FALSE,
                      max_width = NULL, pixel_height = NULL){
  if(interactive_plots){
    w = plotly::ggplotly(p, tooltip = "text", height = pixel_height)
    if(isTRUE(lock_aspect))
      w = plotly::layout(w,
                         yaxis = list(scaleanchor = "x", scaleratio = 1))
    if(isTRUE(legend_below)){
      w = plotly::layout(w,
                         margin = list(b = 170, t = 40),
                         legend = list(orientation = "h",
                                       x = 0.5, xanchor = "center",
                                       y = -0.40, yanchor = "top"))
      # Continuous colorbars live on each trace's `marker.colorbar`, not in
      # the layout legend. Walk the figure data and modify ONLY traces that
      # already have a colorbar (discrete-color traces don't, so they're
      # untouched — avoids the spurious-colorbar bug from earlier).
      for(.i in seq_along(w$x$data)){
        .mk = w$x$data[[.i]]$marker
        if(isTRUE(.mk$showscale) || !is.null(.mk$colorscale)){
          if(is.null(w$x$data[[.i]]$marker$colorbar))
            w$x$data[[.i]]$marker$colorbar = list()
          w$x$data[[.i]]$marker$colorbar$orientation = "h"
          w$x$data[[.i]]$marker$colorbar$x          = 0.5
          w$x$data[[.i]]$marker$colorbar$xanchor    = "center"
          w$x$data[[.i]]$marker$colorbar$y          = -0.30
          w$x$data[[.i]]$marker$colorbar$yanchor    = "top"
          w$x$data[[.i]]$marker$colorbar$len        = 0.7
          w$x$data[[.i]]$marker$colorbar$thickness  = 14
        }
      }
    }
    w = plotly::config(w, displaylogo = FALSE,
                       modeBarButtonsToRemove = c("lasso2d", "select2d"))
    tags = htmltools::as.tags(w, standalone = FALSE)
    deps = htmltools::findDependencies(tags)
    if(length(deps)) knitr::knit_meta_add(list(deps))
    html_str = as.character(tags)
    if(!is.null(max_width))
      html_str = sprintf('<div style="max-width:%s;margin:0 auto;">%s</div>',
                         max_width, html_str)
    cat(html_str)
  } else {
    print(p)
  }
}

# Per-point hover for the box/jitter layers: sample id, y-value (labelled with
# the datatype units) and the primary study factor. Kept here because the
# report's tooltip wording is report-level presentation, not plot logic.
.point_tooltip = function(sid, gsf_val, yval, ylab = value_label){
  sprintf("%s: %s<br>%s: %.3g<br>%s: %s",
          sample_id_head, sid, ylab, yval, global_summary_factor, gsf_val)
}

# plotBoxplot() draws one type at a time; the report shows one or both.
# Every box/bar in the report takes the package qualitative palette, so a
# group keeps the same colour whether it is drawn here, in a heatmap
# annotation or on the PCA.
.compound_plots = function(de_plot, compound, group_by, facet_by){
  types = if(plot_type == "both") c("box", "bar") else plot_type
  out = lapply(types, function(ty)
    plotBoxplot(de_plot, compound, group_by, facet_by = facet_by,
                 type           = ty,
                 error_bar      = plot_error_bar,
                 value_label    = units,
                 sample_id_head = sample_id_head,
                 tooltip_factor = global_summary_factor,
                 yaxis_quant    = yaxis_quant,
                 yaxis_scalar   = yaxis_scalar))
  stats::setNames(out, types)
}
```

# 1. Aim

Statistical analysis of processed `r units` data. Pre-defined comparisons,
correlations, and linear models from the YAML config; multivariate (PCA +
heatmap) inspection per comparison.

> **All per-feature numerical results — comparisons p-values, correlations,
> linear-model coefficients — are written to the stats xlsx (file path
> ends with `_stats.xlsx`) saved alongside this HTML report. Open it in
> Excel to filter / sort the full tables.** This report shows only the
> visualisations and the volcano significant-feature shortlist; the
> long-form tables are intentionally not duplicated here. The workbook's
> first sheet, `key`, defines every column heading it contains.

**Missing-value policy** (which sections impute and which don't):

`r .n_imp = .n_imputed_total(combined_datamatrices); .is_imp = if(exists("values_imputed")) isTRUE(values_imputed) else .n_imp > 0; if(.is_imp) sprintf("This report uses an **imputed** matrix - %s - so the sections below that say *no imputation* include filled-in values%s.", if(.n_imp > 0) sprintf("%d value(s) in it were filled in (\x60n_imputed\x60 in the workbook)", .n_imp) else "how many values were filled in was not recorded", if(isTRUE(report_pars$use_imputed)) " (\x60use_imputed: True\x60)" else ": its input was imputed, e.g. a combine input or a dataset stored before measured values were kept") else "This report uses the values as **measured**, gaps included; only the PCA fills gaps - see its row in the table below."`

| Section | NA handling |
|---|---|
| Comparisons (t-test / Wilcoxon / ANOVA / Kruskal / Tukey) | **No imputation** — raw values used; NAs dropped per test by `t.test()` / `wilcox.test()` / `aov()`. |
| Boxplots (per-compound + by-factor) | **No imputation** — raw values plotted; NAs simply absent from the boxes/jitter. |
| Correlations | **No imputation** — `pairwise.complete.obs` (each pair uses its own complete cases); a pair with fewer than 3 shared samples is left blank. |
| Linear models | `lm()` drops rows with NA in the feature or any predictor (per feature). |
| PCA (global + per-comparison) | `r if(isTRUE(get0("pca_imputed"))) "The values imputed in pre-processing (\x60replace_MVs\x60), as in the xlsx matrix" else sprintf("Imputed per \x60pca.preprocess.impute\x60 (in this config: %s%s)", pca_impute, if(pca_impute %in% c("frac_min", "gaussian")) sprintf(" with frac=%g", pca_impute_frac) else "")`, then transform → centre → scale. |
| Heatmap | After `r heatmap_transform` transform and per-feature z-scoring, values are capped at ±2.5. Missing cells are drawn grey; only for ordering and clustering do they count as the per-feature mean. |
| Volcano | Inherits the comparisons table, so no imputation in the underlying p-values; rows missing log2FC or p are dropped from the plot. |

# 2. Data import

```{r import}
importData     = combined_datamatrices
metadata       = importData$sample_meta
data_matrix    = importData$data
compoundInfo   = importData$variable_meta
compound_names = compoundInfo$Compound

# Biological-sample subset: dropped from sections that should not see blanks/QCs
# (global summary, global PCA). Per-comparison sections do their own subsetting
# via subsetDataset(), so they're unaffected.
if(!is.null(bio_filter_head) && bio_filter_head %in% colnames(metadata)){
  bio_de = subsetDataset(importData,
                     conditions = setNames(list(bio_filter_levels),
                                           bio_filter_head))
  bio_meta   = bio_de$sample_meta
  bio_matrix = bio_de$data
} else {
  bio_de     = importData
  bio_meta   = metadata
  bio_matrix = data_matrix
}

# Heatmap captions name the feature grouping only when it was applied; a
# configured column the data does not have is said so rather than claimed.
.gf_caption = function(col){
  res = .resolve_meta_col(col, as.data.frame(bio_de$variable_meta))
  if(!is.null(res)) sprintf(", features grouped by \x60%s\x60", res)
  else if(!is.null(col))
    sprintf(" (\x60%s\x60 is not a feature_metadata column, so features are not grouped)", col)
  else ""
}

# Match configured feature columns to their stored names once, so a heading
# renamed on import (S-group stored as S.group) works in every heatmap
# section - several of them compare names exactly. A name that does not
# resolve is kept as written, so the sections can say it is missing.
.vm_bio    = as.data.frame(bio_de$variable_meta)
.as_stored = function(col) .resolve_meta_col(col, .vm_bio) %||% col
heatmap_group_features_by             = .as_stored(heatmap_group_features_by)
heatmap_general_cfg$group_features_by = .as_stored(heatmap_general_cfg$group_features_by)
heatmap_topchg_cfg$group_features_by  = .as_stored(heatmap_topchg_cfg$group_features_by)
heatmap_split_cfg$group_features_by   = .as_stored(heatmap_split_cfg$group_features_by)
heatmap_subclass_col                  = .as_stored(heatmap_subclass_col)

# Top-changing caption: per class only when the class column exists.
.tc_caption = function(col){
  if(!is.null(col) && col %in% colnames(.vm_bio))
    sprintf("Top %d features per \x60%s\x60 class.", heatmap_topchg_n, col)
  else if(!is.null(col))
    sprintf("Top %d features overall (\x60%s\x60 is not a feature_metadata column, so features are not grouped).", heatmap_topchg_n, col)
  else sprintf("Top %d features overall.", heatmap_topchg_n)
}

# Fail loud here rather than let downstream sections silently degrade:
# smeta[[factor_name]] on a missing column returns NULL (no error), which
# would just produce an uninformative single-group summary bar/heatmap
# instead of surfacing the YAML typo.
if(!global_summary_factor %in% colnames(metadata))
  stop(sprintf(
    "[stats_report] global_summary_factor='%s' is not a column in sample_metadata. Available columns: %s",
    global_summary_factor, paste(colnames(metadata), collapse = ", ")))

DT::datatable(
  metadata %>%
    dplyr::group_by(.data[[sample_type_head]]) %>%
    dplyr::summarize(`Number of samples` = dplyr::n(), .groups = "drop")
)
```

`r if(!is.null(bio_filter_head)) sprintf("_Biological-sample filter applied to global summary + global PCA: \x60%s\x60 ∈ {%s} → %d samples (of %d)._", bio_filter_head, paste(bio_filter_levels, collapse=", "), nrow(bio_meta), nrow(metadata))`

# 3. Global summary {.tabset}

Mean intensity per compound (biological subset: `r nrow(bio_meta)` of
`r nrow(metadata)` samples), with `r global_summary_factor` levels dodged
side-by-side. Compounds in TargetLynx import order. Plot height auto-scales
with the number of features.

```{r global-helpers, include=FALSE}
.feature_bar = function(d, smeta, factor_name, scale, units){
  feats = colnames(d)
  grp   = smeta[[factor_name]]

  rows = lapply(feats, function(f){
    by_grp = tapply(d[[f]], grp, mean, na.rm = TRUE)
    data.frame(feature = f,
               .grp    = names(by_grp),
               value   = unname(by_grp),
               stringsAsFactors = FALSE)
  })
  m = dplyr::bind_rows(rows)

  if(scale == "log2") m$value = suppressWarnings(log2(m$value + 1))
  if(scale == "sqrt") m$value = suppressWarnings(sqrt(pmax(m$value, 0)))

  # Reverse so first imported feature sits at the top of the y-axis
  m$feature = factor(m$feature, levels = rev(feats))

  xlab = switch(scale,
                raw  = units,
                log2 = paste0("log2(", units, " + 1)"),
                sqrt = paste0("sqrt(", units, ")"))

  ggplot2::ggplot(m, ggplot2::aes(value, feature, fill = .grp)) +
    ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8),
                      width = 0.7) +
    ggplot2::scale_fill_manual(values = .group_fill_colors(m$.grp),
                               na.value = "grey70") +
    ggplot2::labs(x = xlab, y = "Compound", fill = factor_name) +
    ggplot2::theme_classic(base_size = 11) +
    ggplot2::theme(
      axis.text.y        = ggplot2::element_text(size = 6),
      legend.position    = "bottom",
      panel.grid.major.x = ggplot2::element_line(color = "grey90", linewidth = 0.3)
    )
}
```

## 3.1 Raw (`r units`) {-}

```{r global-feature-raw, fig.width=7*fig_scale, fig.height=max(8, ncol(bio_matrix)*0.18)*fig_scale}
print(.feature_bar(bio_matrix, bio_meta, global_summary_factor, "raw", units))
```

## 3.2 log2 (`r units`) {-}

```{r global-feature-log2, fig.width=7*fig_scale, fig.height=max(8, ncol(bio_matrix)*0.18)*fig_scale}
print(.feature_bar(bio_matrix, bio_meta, global_summary_factor, "log2", units))
```

# 4. Comparisons

```{r comparisons-summary, results="asis"}
if(length(comparisons) == 0){
  cat("_No comparisons defined in YAML._\n")
} else {
  cat("Configured comparisons:\n\n")
  for(comp in comparisons){
    sub_str = if(is.null(comp$subset)) "(all samples)" else
      paste(sprintf("%s = %s", names(comp$subset), unlist(comp$subset)),
            collapse = ", ")
    cat(sprintf("- **%s** — subset: `%s`; compare `%s`: %s\n",
                comp$name, sub_str, comp$compare$factor,
                paste(comp$compare$levels, collapse = " vs ")))
  }
}
```

**Methods.** For each comparison the test method is auto-selected by group
count: 2 levels → t-test + Wilcoxon; ≥3 levels → ANOVA + Kruskal-Wallis,
plus Tukey HSD post-hoc per pair. P-values are Benjamini-Hochberg adjusted
across **all** comparisons × methods × features.

> **Full p-value tables (per comparison + per method, all features) are
> in the `stats` tab of the stats xlsx** written by `runMRManalyzeR()`
> alongside the rendered HTML reports. The same xlsx also carries the
> `correlations` and `linear_models` tabs.

## 4.1 Per-compound `r switch(plot_type, box="boxplot", bar="barplot", both="boxplot + barplot")` (faceted across comparisons) {-}

`r if(!plot_boxplot_per_comparison) "_Skipped — set_ \x60plot_boxplot_per_compound: True\x60 _in YAML to enable._"`

```{r comp-boxplots, results="asis", eval=plot_boxplot_per_comparison && length(comparisons) > 0, fig.width=10*fig_scale, fig.height=6*fig_scale}
# Build one long data frame across all comparisons, then facet.
# Combine user `subset:` with the comparison level restriction in one
# subsetDataset() call (avoid post-hoc DE mutation which trips SE's assay check).
long_rows = list()
for(comp in comparisons){
  cond = comp$subset
  cond = if(is.null(cond)) list() else as.list(cond)
  cond[[comp$compare$factor]] = comp$compare$levels
  de_sub = subsetDataset(importData, conditions = cond)
  if(nrow(de_sub$sample_meta) == 0) next

  d = de_sub$data
  .sm = de_sub$sample_meta
  d$.sample_id  = if(sample_id_head %in% colnames(.sm))
                    as.character(.sm[[sample_id_head]]) else rownames(d)
  d$.group      = factor(de_sub$sample_meta[[comp$compare$factor]],
                         levels = comp$compare$levels)
  d$.comparison = comp$name
  d$.gsf        = if(global_summary_factor %in% colnames(.sm))
                    as.character(.sm[[global_summary_factor]]) else NA_character_
  long_rows[[length(long_rows) + 1]] = d
}
long = dplyr::bind_rows(long_rows)

make_safe_id = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
safe_ids = vapply(compound_names, make_safe_id, character(1))

is_detected = vapply(compound_names, function(cmp){
  cc = which(colnames(long) == cmp)
  length(cc) > 0 && !all(is.na(unlist(long[, cc])))
}, logical(1))
detected_compounds = compound_names[is_detected]
detected_ids       = safe_ids[is_detected]

# Alphabetise dropdown order (T3.5-7); plot blocks below iterate in import order.
.alpha_idx        = order(detected_compounds)
detected_alpha    = detected_compounds[.alpha_idx]
detected_ids_alpha = detected_ids[.alpha_idx]

cat('
<label for="boxplot_select" style="font-weight:600;">Select compound:</label>
<select id="boxplot_select" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showBoxplot(this.value)">
')
for(i in seq_along(detected_alpha)){
  sel = if(i == 1) " selected" else ""
  cat(sprintf('<option value="%s"%s>%s</option>\n',
              detected_ids_alpha[i], sel, detected_alpha[i]))
}
cat('</select>
<script>
function showBoxplot(id){
  var plots = document.querySelectorAll(".cmp_boxplot_block");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("cmp_boxplot_" + id);
  if (el) {
    el.style.display = "block";
    if (window.Plotly){
      var pls = el.querySelectorAll(".plotly-graph-div");
      for (var j=0; j<pls.length; j++) Plotly.Plots.resize(pls[j]);
    }
  }
}
</script>
')

# One DatasetExperiment over the long frame: rows are sample x comparison
# pseudo-injections, which is exactly what the faceted panels show.
.long_meta = long[, c(".group", ".comparison"), drop = FALSE]
# Name these as the real metadata columns so the tooltip labels read
# "Sample_ID: ..." / "Treatment: ..." rather than the internal names.
.long_meta[[sample_id_head]]        = long$.sample_id
.long_meta[[global_summary_factor]] = long$.gsf

.long_de = struct::DatasetExperiment(
  data          = long[, detected_compounds, drop = FALSE],
  sample_meta   = .long_meta,
  variable_meta = data.frame(Compound = detected_compounds,
                             row.names = detected_compounds))

for(i in seq_along(detected_compounds)){
  compound = detected_compounds[i]
  cid      = detected_ids[i]
  plots = .compound_plots(.long_de, compound, ".group",
                          facet_by = ".comparison")

  # Show alphabetically-first detected compound by default (matches dropdown)
  display = if(detected_ids[i] == detected_ids_alpha[1]) "block" else "none"
  cat(sprintf('<div class="cmp_boxplot_block" id="cmp_boxplot_%s" style="display:%s;">\n',
              cid, display))
  # When plot_type is "both" we get two separate figures (box first, bar second)
  for(pname in names(plots)){
    if(plot_type == "both")
      cat(sprintf('<div style="margin-top:10px;"><b>%s</b></div>\n',
                  if(pname == "box") "Boxplot" else "Barplot"))
    .show_plot(plots[[pname]])
  }
  cat("</div>\n")
}
```

## 4.2 Per-compound `r switch(plot_type, box="boxplot", bar="barplot", both="boxplot + barplot")` by factor(s) {-}

`r if(!boxplot_enabled) "_Skipped — set_ \x60boxplot.enabled: True\x60 _in YAML to enable._"`

`r if(boxplot_enabled) sprintf("Per-compound boxplot grouped by_ \x60%s\x60_. Display mode:_ \x60%s\x60_._", paste(boxplot_factors, collapse=", "), boxplot_display)`

```{r factor-boxplots, results="asis", eval=boxplot_enabled && length(boxplot_factors) > 0, fig.width=10*fig_scale, fig.height=6*fig_scale}
# Filter to requested sample types
keep = metadata[[boxplot_head]] %in% boxplot_levels
plot_meta = metadata[keep, , drop = FALSE]
plot_data = data_matrix[keep, , drop = FALSE]

# Primary study factor per row (for point tooltips), aligned to plot_data rows.
gsf_fb = if(global_summary_factor %in% colnames(plot_meta))
           as.character(plot_meta[[global_summary_factor]]) else NA_character_

missing_bf = setdiff(boxplot_factors, colnames(plot_meta))
if(length(missing_bf))
  stop(sprintf(
    "[stats_report] boxplot.factors contains column(s) not in sample_metadata: %s. Available columns: %s",
    paste(missing_bf, collapse = ", "), paste(colnames(plot_meta), collapse = ", ")))

# Detect compounds with any data
is_detected_fb = vapply(compound_names, function(cmp){
  cc = which(colnames(plot_data) == cmp)
  length(cc) > 0 && !all(is.na(unlist(plot_data[, cc])))
}, logical(1))
detected_fb     = compound_names[is_detected_fb]
make_safe_id_fb = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
detected_ids_fb = vapply(detected_fb, make_safe_id_fb, character(1))

# Build per-tab grouping vectors
tab_groupings = list()
if(boxplot_display == "combined" || length(boxplot_factors) == 1){
  lab = paste(boxplot_factors, collapse = " × ")
  tab_groupings[[lab]] = do.call(paste, c(plot_meta[boxplot_factors], sep = "_"))
} else {
  for(f in boxplot_factors)
    tab_groupings[[f]] = as.character(plot_meta[[f]])
  if(length(boxplot_factors) >= 2){
    lab = paste(boxplot_factors, collapse = " × ")
    tab_groupings[[lab]] = do.call(paste, c(plot_meta[boxplot_factors], sep = "_"))
  }
}

cat('\n### {.tabset .tabset-fade}\n\n')

for(tab_name in names(tab_groupings)){
  cat(sprintf('\n#### %s {-}\n\n', tab_name))

  grouping = tab_groupings[[tab_name]]
  tab_safe = make_safe_id_fb(tab_name)

  # Alphabetised dropdown (T3.5-7); plot blocks below in import order.
  alpha_idx_fb       = order(detected_fb)
  detected_alpha_fb  = detected_fb[alpha_idx_fb]
  detected_ids_alpha_fb = detected_ids_fb[alpha_idx_fb]

  cat(sprintf('
<label for="fbox_select_%s" style="font-weight:600;">Select compound:</label>
<select id="fbox_select_%s" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showFactorBoxplot_%s(this.value)">
', tab_safe, tab_safe, tab_safe))
  for(i in seq_along(detected_alpha_fb)){
    sel = if(i == 1) " selected" else ""
    cat(sprintf('<option value="%s"%s>%s</option>\n',
                detected_ids_alpha_fb[i], sel, detected_alpha_fb[i]))
  }
  cat(sprintf('</select>
<script>
function showFactorBoxplot_%s(id){
  var plots = document.querySelectorAll(".fbox_block_%s");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("fbox_%s_" + id);
  if (el) {
    el.style.display = "block";
    if (window.Plotly){
      var pls = el.querySelectorAll(".plotly-graph-div");
      for (var j=0; j<pls.length; j++) Plotly.Plots.resize(pls[j]);
    }
  }
}
</script>
', tab_safe, tab_safe, tab_safe))

  .tab_meta = plot_meta
  .tab_meta$.group = grouping
  .tab_de = struct::DatasetExperiment(
    data          = plot_data[, detected_fb, drop = FALSE],
    sample_meta   = .tab_meta,
    variable_meta = data.frame(Compound = detected_fb,
                               row.names = detected_fb))

  for(i in seq_along(detected_fb)){
    compound = detected_fb[i]
    cid      = detected_ids_fb[i]
    plots = .compound_plots(.tab_de, compound, ".group",
                            facet_by = NULL)

    # Show alphabetically-first compound by default (matches dropdown)
    display = if(detected_ids_fb[i] == detected_ids_alpha_fb[1]) "block" else "none"
    cat(sprintf('<div class="fbox_block_%s" id="fbox_%s_%s" style="display:%s;">\n',
                tab_safe, tab_safe, cid, display))
    for(pname in names(plots)){
      if(plot_type == "both")
        cat(sprintf('<div style="margin-top:10px;"><b>%s</b></div>\n',
                    if(pname == "box") "Boxplot" else "Barplot"))
      .show_plot(plots[[pname]])
    }
    cat("</div>\n")
  }
}
```

## 4.3 PCA per comparison

`r if(!pca_include) "_Skipped — set_ \x60pca.enabled: True\x60 _in YAML to enable._"`

For each comparison: filter samples by the comparison's `subset:`, run PCA on
that slice, colour points by the comparison's `compare.factor`. The first
tab is a **global** PCA across the biological subset, coloured by
`r global_summary_factor`.

**Components plotted:** PC`r pca_pc_x` × PC`r pca_pc_y`
(set via `stats_report.pca.components: [x, y]`).

**Preprocessing pipeline** (configured in `stats_report.pca.preprocess`; per-run actual steps are echoed inside each tab):

| Step | Setting | Value |
|---|---|---|
| Feature filter | `feat_na_max` | `r if(pca_feat_na_max < 1) sprintf("drop features with > %g%% NA", pca_feat_na_max*100) else "none (only all-NA features removed)"` |
| Sample filter | `sample_na_max` | `r if(pca_sample_na_max < 1) sprintf("drop samples with > %g%% NA", pca_sample_na_max*100) else "none"` |
| Imputation | `impute` | `r pca_impute``r if(pca_impute %in% c("frac_min", "gaussian")) sprintf(" (frac = %g)", pca_impute_frac)` |
| Transform | `transform` | `r pca_transform` |
| Centre | `center` | `r pca_center` |
| Scale | `scale` | `r pca_scale` |

```{r pca-helpers, include=FALSE}
# Local helper: outlier flag on chosen PCs
.pca_outliers = function(x, y){
  zx = scale(x)[, 1]; zy = scale(y)[, 1]
  abs(zx) > 2.5 | abs(zy) > 2.5
}

# Build the scores ggplot object. Does NOT print — caller decides.
.pca_scores_plot = function(res_pca, smeta, color_factor,
                            pc_x, pc_y, label_mode, label_factor,
                            ellipse_mode, ellipse_conf){
  pr = res_pca$pr
  if(is.null(pr)) return(NULL)

  # Validate / fall back component selection
  pc_max = ncol(pr$x)
  if(pc_x > pc_max || pc_y > pc_max){
    pc_x = 1; pc_y = min(2, pc_max)
  }
  vexp = (pr$sdev^2) / sum(pr$sdev^2) * 100
  pcx_lab = sprintf("PC%d", pc_x); pcy_lab = sprintf("PC%d", pc_y)

  scores = data.frame(
    pcx   = pr$x[, pc_x],
    pcy   = pr$x[, pc_y],
    group = smeta[[color_factor]],
    stringsAsFactors = FALSE
  )
  scores$label = if(!is.null(label_factor) && label_factor %in% colnames(smeta))
                   as.character(smeta[[label_factor]]) else rownames(scores)
  scores$tooltip = sprintf("%s: %s<br>%s: %s<br>PC%d: %.3g<br>PC%d: %.3g",
                           label_factor, scores$label,
                           color_factor, as.character(scores$group),
                           pc_x, scores$pcx, pc_y, scores$pcy)
  is_num = is.numeric(scores$group)

  p = ggplot2::ggplot(scores, ggplot2::aes(pcx, pcy, color = group, text = tooltip)) +
        ggplot2::geom_point(size = 3, alpha = 0.85) +
        ggplot2::labs(x = sprintf("%s (%.1f%%)", pcx_lab, vexp[pc_x]),
                      y = sprintf("%s (%.1f%%)", pcy_lab, vexp[pc_y]),
                      color = color_factor) +
        ggplot2::theme_classic(base_size = 13) +
        ggplot2::theme(
          legend.position   = "bottom",
          legend.box.margin = ggplot2::margin(t = 12),
          axis.title.x      = ggplot2::element_text(margin = ggplot2::margin(t = 6, b = 0)),
          plot.margin       = ggplot2::margin(4, 6, 4, 6)
        )
  p = if(is_num) p + ggplot2::scale_color_viridis_c()
      else       p + ggplot2::scale_color_manual(
                       values = .group_fill_colors(scores$group),
                       na.value = "grey60")
  if(ellipse_mode == "group" && !is_num && length(unique(scores$group)) > 1)
    p = p + ggplot2::stat_ellipse(ggplot2::aes(group = group),
                                  type = "norm", level = ellipse_conf,
                                  linewidth = 0.4)
  if(label_mode == "all"){
    p = p + ggplot2::geom_text(ggplot2::aes(label = label),
                               size = 2.5, vjust = -1, show.legend = FALSE)
  } else if(label_mode == "outliers"){
    out_idx = .pca_outliers(scores$pcx, scores$pcy)
    if(any(out_idx))
      p = p + ggplot2::geom_text(data = scores[out_idx, ],
                                 ggplot2::aes(label = label),
                                 size = 2.5, vjust = -1, show.legend = FALSE)
  }
  p
}

# Loadings are drawn by plotLoadings(); this wrapper only carries the
# report's component selection and colour column.
.pca_loadings_plot = function(res_pca, vmeta, pc_x, pc_y, loadings_color_by){
  plotLoadings(res_pca, vmeta, components = c(pc_x, pc_y),
                colour_by = loadings_color_by)
}

# Print preprocessing trace + the matrix-size note (used in scores tab).
.pca_preprocess_trace = function(res_pca){
  cat("**Preprocessing pipeline applied:**\n\n")
  for(s in res_pca$steps) cat(sprintf("- %s\n", s))
  cat(sprintf("\n_Final matrix: %d samples × %d features._\n\n",
              res_pca$n_samples, res_pca$n_features))
  pr = res_pca$pr
  if(is.null(pr))
    cat(sprintf("_PCA could not be fit (n_samples=%d, n_features=%d)._\n\n",
                res_pca$n_samples, res_pca$n_features))
}
```

```{r comp-pca-compute, include=FALSE, eval=pca_include}
# Pre-compute every PCA (Global + per-comparison) into a list shared by the
# two rendering chunks below. Each entry holds the PCA result, the metadata
# slices, the colour factor, and a tab heading. We compute once and render
# twice (scores at small canvas, loadings at tall canvas) to give the scores
# plot a sensible aspect ratio independent of feature count.
.pca_runs = list()

# Helper: run the PCA pipeline AND realign sample_meta to the samples that
# actually survived `sample_na_max`. Without this, scores ↔ smeta mismatch
# when sample_na_max < 1 drops any rows.
.pca_run_one = function(de){
  M  = as.matrix(de$data)
  # Same samples and compounds, values from the exported matrix: what
  # replace_MVs imputed, so runPCA() fills only what is still missing.
  P = get0("pca_datamatrices")
  if(!is.null(P)){
    P = as.matrix(as.data.frame(P$data))
    if(all(rownames(M) %in% rownames(P)) && all(colnames(M) %in% colnames(P)))
      M = P[rownames(M), colnames(M), drop = FALSE]
  }
  sm = de$sample_meta
  # Anchor rownames so we can subset sm after the pipeline drops rows.
  rn = if(!is.null(rownames(M)) && !anyDuplicated(rownames(M))) rownames(M)
       else paste0("s", seq_len(nrow(M)))
  rownames(M) = rn
  rownames(sm) = rn

  res = runPCA(M,
                         impute        = pca_impute,
                         impute_frac   = pca_impute_frac,
                         seed          = pca_seed,
                         transform     = pca_transform,
                         center        = pca_center,
                         scale         = pca_scale,
                         feat_na_max   = pca_feat_na_max,
                         sample_na_max = pca_sample_na_max)
  if(!is.null(res$pr))
    sm = sm[rownames(res$pr$x), , drop = FALSE]

  list(res_pca = res, smeta = sm, vmeta = de$variable_meta)
}

# Global tab
pca_global_de = if(length(pca_global_subset)){
  subsetDataset(bio_de, conditions = as.list(pca_global_subset))
} else {
  bio_de
}
global_subset_lbl = if(length(pca_global_subset)){
  sprintf(" — subset: %s",
          paste(sprintf("%s=%s", names(pca_global_subset),
                        unlist(pca_global_subset)), collapse = ", "))
} else { "" }
.gbl = .pca_run_one(pca_global_de)
.pca_runs[["Global"]] = list(
  tab_heading  = sprintf("Global (%d samples; coloured by %s)%s",
                         nrow(.gbl$smeta),
                         global_summary_factor, global_subset_lbl),
  res_pca      = .gbl$res_pca,
  smeta        = .gbl$smeta,
  vmeta        = .gbl$vmeta,
  color_factor = global_summary_factor,
  insufficient = FALSE
)

# Per-comparison tabs
for(comp in comparisons){
  cond = comp$subset
  cond = if(is.null(cond)) list() else as.list(cond)
  cond[[comp$compare$factor]] = comp$compare$levels
  de_sub = subsetDataset(importData, conditions = cond)

  if(nrow(de_sub$data) < 3 || ncol(de_sub$data) < 2){
    .pca_runs[[comp$name]] = list(tab_heading = comp$name, insufficient = TRUE)
    next
  }
  .one = .pca_run_one(de_sub)
  .pca_runs[[comp$name]] = list(
    tab_heading  = comp$name,
    res_pca      = .one$res_pca,
    smeta        = .one$smeta,
    vmeta        = .one$vmeta,
    color_factor = comp$compare$factor,
    insufficient = FALSE
  )
}

# Cap the loadings canvas height by the largest feature count across runs.
.pca_max_feats = max(c(0, vapply(.pca_runs, function(r){
  if(isTRUE(r$insufficient) || is.null(r$res_pca$pr)) 0L
  else nrow(r$res_pca$pr$rotation)
}, integer(1))))
```

### Scores {.tabset .tabset-fade}

```{r comp-pca-scores, results="asis", eval=pca_include, fig.width=4.5*fig_scale, fig.height=4.5*fig_scale, out.width="450px", out.height="auto"}
for(nm in names(.pca_runs)){
  r = .pca_runs[[nm]]
  cat(sprintf("\n\n#### %s {-}\n\n", r$tab_heading))
  if(isTRUE(r$insufficient)){ cat("_Insufficient data._\n\n"); next }
  .pca_preprocess_trace(r$res_pca)
  p_sc = .pca_scores_plot(r$res_pca, r$smeta, r$color_factor,
                          pca_pc_x, pca_pc_y,
                          pca_label, pca_label_factor,
                          pca_ellipse, pca_ellipse_conf)
  if(!is.null(p_sc)){
    .show_plot(p_sc, lock_aspect = TRUE, legend_below = TRUE,
               max_width = "650px", pixel_height = 650)
    cat("\n\n")
  }
}
```

### Loadings {.tabset .tabset-fade}

```{r comp-pca-loadings, results="asis", eval=pca_include && pca_loadings_show, fig.width=10*fig_scale, fig.height=max(7, (if(exists(".pca_max_feats")) .pca_max_feats else 0) * 0.18) * fig_scale}
for(nm in names(.pca_runs)){
  r = .pca_runs[[nm]]
  cat(sprintf("\n\n#### %s {-}\n\n", r$tab_heading))
  if(isTRUE(r$insufficient) || is.null(r$res_pca$pr)){
    cat("_No loadings available._\n\n"); next
  }
  ld_col = .resolve_meta_col(pca_loadings_color_by, as.data.frame(r$vmeta))
  if(!is.null(ld_col)){
    cat(sprintf("**Loadings** _(colour =_ \x60%s\x60_; bars grouped by colour, sorted within each group)_\n\n",
                ld_col))
  } else if(!is.null(pca_loadings_color_by)){
    cat(sprintf("**Loadings** _(\x60%s\x60 is not a feature_metadata column, so the bars are not coloured)_\n\n",
                pca_loadings_color_by))
  } else {
    cat("**Loadings**\n\n")
  }
  p_ld = .pca_loadings_plot(r$res_pca, r$vmeta, pca_pc_x, pca_pc_y,
                            pca_loadings_color_by)
  if(!is.null(p_ld)){ print(p_ld); cat("\n\n") }
}
```

## 4.4 Volcano per comparison {.tabset .tabset-fade}

`r if(!volcano_include) "_Skipped — set_ \x60volcano.enabled: True\x60 _in YAML to enable._"`

**Methods.** One volcano per 2-group comparison directly from the t-test
rows. For ≥3-group comparisons we plot one volcano per Tukey HSD pair.
X = log2 fold-change (mean_g2 / mean_g1); Y = -log10 of `r if(volcano_use_adj) "BH-adjusted" else "raw"` p
(toggle via `stats_report.volcano.use_adjusted`).
Top `r volcano_top_n` features by significance are labelled.
Significance threshold dashed line at p `r if(volcano_use_adj) "(BH-adjusted)" else "(raw)"` = `r sig_threshold`.

```{r comp-volcano, results="asis", eval=volcano_include && !is.null(stats_tables) && nrow(stats_tables$stats) > 0, fig.width=5.5*fig_scale, fig.height=5*fig_scale, out.width="600px"}
# Volcano sources:
#   2-group comparisons -> the single test row (welch / student / wilcoxon)
#   3+-group comparisons -> the post-hoc rows (one volcano per pair)
# Global tests (anova / kruskal) have no fold change and are excluded.
.volc_p_col = if(volcano_use_adj) "p_adj" else "p_value"
.volc_p_lbl = if(volcano_use_adj) sprintf("BH-adjusted p (%s)", p_adjust) else "raw p"

volc_data = stats_tables$stats %>%
  dplyr::filter(!is.na(log2FC), !is.na(.data[[.volc_p_col]]))

if(nrow(volc_data) == 0){
  cat("_No volcano-eligible rows._\n")
} else {
  for(comp_name in unique(volc_data$comparison)){
    sub = volc_data %>% dplyr::filter(comparison == comp_name)
    .posthoc = c("tukey", "pairwise_wilcoxon")
    pair_groups = if(any(sub$method %in% .posthoc))
      unique(sub %>% dplyr::filter(method %in% .posthoc) %>%
               dplyr::transmute(pair = paste(group1, "vs", group2)) %>%
               dplyr::pull(pair))
    else NA

    for(pair_lbl in (if(is.na(pair_groups[1])) NA else pair_groups)){
      this = if(is.na(pair_lbl)) sub else
        sub %>% dplyr::filter(method %in% .posthoc,
                              paste(group1, "vs", group2) == pair_lbl)

      cat(sprintf("\n### %s%s {-}\n\n", comp_name,
                  if(is.na(pair_lbl)) "" else paste0(" — ", pair_lbl)))

      p = plotVolcano(this, comparison = comp_name,
                       pair          = if(is.na(pair_lbl)) NULL else pair_lbl,
                       use_adjusted  = volcano_use_adj,
                       sig_threshold = sig_threshold,
                       top_n_label   = volcano_top_n)

      sig_rows = this %>% dplyr::filter(.data[[.volc_p_col]] < sig_threshold) %>%
                          dplyr::arrange(.data[[.volc_p_col]])

      .show_plot(p, max_width = "650px")
      cat("\n\n")

      if(nrow(sig_rows) > 0){
        sig_tbl = sig_rows %>%
          dplyr::select(feature, log2FC,
                        dplyr::any_of(c("mean_g1", "mean_g2")),
                        dplyr::all_of(.volc_p_col)) %>%
          dplyr::mutate(dplyr::across(where(is.numeric), ~ signif(., 4)))
        # kable renders reliably in results="asis" chunks that also emit plots.
        # DT htmlwidgets are not captured correctly when mixed with ggplot
        # device output in the same asis chunk.
        cap_html = sprintf(
          '<p style="font-size:13px;margin-top:8px;"><b>%d significant feature(s)</b> &mdash; %s &lt; %.3g</p>',
          nrow(sig_tbl), .volc_p_lbl, sig_threshold)
        tbl_html = knitr::kable(
          sig_tbl, format = "html", row.names = FALSE,
          table.attr = 'class="table table-sm table-striped table-bordered" style="font-size:13px;width:auto;"')
        cat(cap_html, '\n<div style="overflow-x:auto;max-height:340px;overflow-y:auto;">\n',
            tbl_html, '\n</div>\n\n', sep = "")
      }
    }
  }
}
```

## 4.5 Heatmap — general (all features) {.tabset .tabset-fade}

`r if(!heatmap_include) "_Skipped — set_ \x60heatmap.enabled: True\x60 _in YAML to enable._"`

`r if(heatmap_include && !heatmap_general_on) "_Skipped — set_ \x60heatmap.general.enabled: True\x60 _in YAML to enable._"`

`r if(heatmap_include && heatmap_general_on) sprintf("All features retained. Cells show per-feature %s of %s; samples coloured by \x60%s\x60%s. Scope: **%s** (Global and/or one tab per comparison). Feature labels: %s.", heatmap_general_cfg$scale, heatmap_general_cfg$transform, heatmap_general_cfg$color_samples_by, .gf_caption(heatmap_general_cfg$group_features_by), heatmap_general_scope, heatmap_general_cfg$feature_labels)`

```{r heatmap, eval=heatmap_include, fig.width=max(10, ncol(bio_matrix)*0.18)*fig_scale, fig.height=max(5, nrow(bio_matrix)*0.22 + 2)*fig_scale}
# Shared heatmap renderer — called for every scope target across 4.5/4.6/4.7.
# M_raw   : samples × features numeric matrix (raw / pre-transform)
# smeta   : sample metadata data.frame, rows aligned to M_raw
# vmeta   : variable metadata data.frame with a 'Compound' column
# title   : plot title string
# cfg     : resolved render settings for the calling section (from .resolve_hm)
#           — transform / scale / color_samples_by / cluster_rows /
#           cluster_within / orientation / group_features_by / feature_labels.
# group_by / feature_labels: explicit overrides (default to the cfg values);
#           the combined split-by-class heatmap overrides group_by to the
#           subclass column.
# Thin wrapper over plotHeatmap(): the scope/target machinery below hands us
# a matrix plus its metadata, so rewrap them as a DatasetExperiment, map the
# resolved cfg onto the function's arguments, and print.
.draw_heatmap = function(M_raw, smeta, vmeta, title, cfg = heatmap_general_cfg,
                         group_by = cfg$group_features_by,
                         feature_labels = cfg$feature_labels,
                         canvas = NULL){
  # 4.6 / 4.7 hand us a feature SUBSET but the full variable_meta, so align
  # them: a DatasetExperiment requires variable_meta rows to match the data
  # columns exactly, one row per feature, in order.
  Mdf = as.data.frame(M_raw)
  vm  = as.data.frame(vmeta)
  sm  = as.data.frame(smeta)

  idx = match(colnames(Mdf), vm$Compound)
  Mdf = Mdf[, !is.na(idx), drop = FALSE]
  vm  = vm[idx[!is.na(idx)], , drop = FALSE]
  if(ncol(Mdf) == 0){
    cat("_No features with metadata to draw._
"); return(invisible(NULL))
  }
  rownames(vm) = vm$Compound
  rownames(sm) = rownames(Mdf)

  # Size the cells to this target's own dimensions against the shared canvas.
  flip = identical(cfg$orientation, "samples_x") ||
         (identical(cfg$orientation, "auto") && nrow(Mdf) > ncol(Mdf))
  n_x  = if(flip) nrow(Mdf) else ncol(Mdf)
  n_y  = if(flip) ncol(Mdf) else nrow(Mdf)
  cell = if(is.null(canvas)) cfg$cell_size
         else .hm_cell_for(n_x, n_y, canvas$w, canvas$h, cfg)

  de_hm = tryCatch(
    struct::DatasetExperiment(data = Mdf, sample_meta = sm,
                              variable_meta = vm),
    error = function(e) NULL)
  if(is.null(de_hm)){
    cat("_Could not assemble heatmap dataset._
"); return(invisible(NULL))
  }

  # One style per heatmap, never both. cluster_rows, cluster_within_groups,
  # orientation and cell_size are properties of the pheatmap grob with no
  # counterpart in the ggplot renderer, and `scale` has none either - the
  # group-mean style always scales per compound, since averaging unscaled
  # values across compounds spanning orders of magnitude has no useful reading.
  # Those settings are ignored under group_means rather than approximated.
  style = .pick_style(cfg$style,
                      length(unique(sm[[cfg$color_samples_by]])))
  p = if(identical(style, "group_means"))
    plotGroupHeatmap(
      de_hm,
      color_samples_by  = cfg$color_samples_by,
      title             = title,
      group_features_by = group_by,
      transform         = cfg$transform,
      alpha_floor       = cfg$alpha_floor,
      feature_labels    = feature_labels,
      sample_id_head    = sample_id_head)
  else
    plotHeatmap(
      de_hm,
      title                 = title,
      color_samples_by      = cfg$color_samples_by,
      group_features_by     = group_by,
      transform             = cfg$transform,
      scale                 = cfg$scale,
      cluster_rows          = cfg$cluster_rows,
      cluster_within_groups = cfg$cluster_within,
      orientation           = cfg$orientation,
      feature_labels        = feature_labels,
      sample_id_head        = sample_id_head,
      cell_size             = cell)

  if(is.null(p)){
    cat("_Insufficient data for heatmap._
"); return(invisible(NULL))
  }
  print(p)
  invisible(NULL)
}

# Build the render targets for a scope selector, shared by 4.5/4.6/4.7. Each
# target is list(label, M = samples x features matrix, smeta, vmeta, comp) —
# comp is NULL for the Global target, else the comparison spec. "comparisons"
# and "all" expand the pre-defined comparisons; empty subsets are dropped.
.heatmap_targets = function(scope){
  g = list(list(label = "Global", M = bio_matrix, smeta = bio_meta,
                vmeta = bio_de$variable_meta, comp = NULL))
  cmps = list()
  if(scope %in% c("comparisons", "all"))
    cmps = Filter(Negate(is.null), lapply(comparisons, function(comp){
      cond = if(is.null(comp$subset)) list() else as.list(comp$subset)
      cond[[comp$compare$factor]] = comp$compare$levels
      de = subsetDataset(importData, conditions = cond)
      if(nrow(de$sample_meta) == 0) return(NULL)
      list(label = comp$name, M = de$data, smeta = de$sample_meta,
           vmeta = de$variable_meta, comp = comp)
    }))
  switch(scope, global = g, comparisons = cmps, all = c(g, cmps))
}

# Option 1: keep only the N most-changing features per class for a comparison.
# Ranks each feature by the chosen stats metric (p_adj / p_value: smaller is
# better; log2fc: larger |value| is better), takes the top N within each
# `group_features_by` class. Falls back to all features if top_changing is off
# or the comparison has no stats.
.top_changing = function(comp_name, feats, vmeta, group_by = heatmap_group_features_by){
  if(!heatmap_topchg_on || is.null(stats_tables) ||
     is.null(stats_tables$stats) || nrow(stats_tables$stats) == 0) return(feats)
  s = stats_tables$stats[stats_tables$stats$comparison == comp_name, , drop = FALSE]
  metric_col = switch(heatmap_topchg_rank,
                      log2fc = "log2FC", p_value = "p_value", "p_adj")
  if(nrow(s) == 0 || !metric_col %in% colnames(s)) return(feats)

  use_abs = metric_col == "log2FC"          # log2FC ranks by magnitude
  agg = tapply(s[[metric_col]], s$feature, function(v){
    v = v[is.finite(v)]
    if(!length(v)) return(NA_real_)
    if(use_abs) max(abs(v)) else min(v)     # best (biggest FC / smallest p)
  })
  score = agg[match(feats, names(agg))]     # aligned to feats; NA if no stats
  grp = if(!is.null(group_by) && group_by %in% colnames(vmeta))
          as.character(vmeta[[group_by]][match(feats, vmeta$Compound)]) else rep("all", length(feats))
  grp[is.na(grp)] = "NA"

  keep = unlist(lapply(unique(grp), function(g){
    idx = which(grp == g)                   # positions in feats for this class
    ord = order(score[idx], decreasing = use_abs, na.last = NA)  # drops NA-score feats
    feats[idx][ord][seq_len(min(heatmap_topchg_n, length(ord)))]
  }), use.names = FALSE)
  if(length(keep)) keep else feats
}

# Global-tab analogue of .top_changing when there is no single comparison to
# rank by: score each feature by how strongly it separates the `color_by` groups
# (the section's color_samples_by / primary study factor). On the transformed
# values, z-score each feature across samples, then score = variance of the
# per-group means; keep the top-N per `group_by` class. Falls back to all feats
# if the factor is absent or has < 2 usable groups.
.top_changing_global = function(feats, M, smeta, vmeta, color_by, group_by, transform){
  if(!heatmap_topchg_on) return(feats)
  if(is.null(color_by) || !color_by %in% colnames(smeta)) return(feats)
  g_all = as.character(smeta[[color_by]]); ok = !is.na(g_all)
  if(length(unique(g_all[ok])) < 2) return(feats)

  X = as.matrix(M[, feats, drop = FALSE])
  if(identical(transform, "log2")){ X[X < 0 & !is.na(X)] = 0; X = log2(X + 1) }
  else if(identical(transform, "sqrt")){ X[X < 0 & !is.na(X)] = 0; X = sqrt(X) }
  X = scale(X)[ok, , drop = FALSE]; g = g_all[ok]     # per-feature z, keep labelled rows
  score = apply(X, 2, function(col){
    m = tapply(col, g, mean, na.rm = TRUE)            # per-group means
    stats::var(as.numeric(m), na.rm = TRUE)           # spread across groups
  })
  names(score) = feats

  cls = if(!is.null(group_by) && group_by %in% colnames(vmeta))
          as.character(vmeta[[group_by]][match(feats, vmeta$Compound)]) else rep("all", length(feats))
  cls[is.na(cls)] = "NA"
  keep = unlist(lapply(unique(cls), function(cc){
    idx = which(cls == cc)
    ord = order(score[idx], decreasing = TRUE, na.last = NA)
    feats[idx][ord][seq_len(min(heatmap_topchg_n, length(ord)))]
  }), use.names = FALSE)
  if(length(keep)) keep else feats
}

# Figure-canvas sizing shared by 4.5/4.6/4.7. Each asis section renders many
# heatmaps into ONE knitr canvas (single fig size per chunk), so size to the
# largest target using that block's orientation. .hm_scope_nrow = max sample
# count across a scope's targets (used for the height/width formula).
.hm_fig_dims = function(max_nrow, n_feat, orientation, fig_scale = 1,
                        cell_size = heatmap_cell_size,
                        max_w = 18, max_h = 22,
                        min_cell = 2, max_cell = 18,
                        max_aspect = heatmap_cell_aspect){
  flip = orientation == "samples_x" || (orientation == "auto" && max_nrow > n_feat)
  n_x = if(flip) max_nrow else n_feat     # cells along the x axis
  n_y = if(flip) n_feat   else max_nrow   # cells along the y axis

  # Fixed allowance (inches) for axis labels, the annotation bars and legend.
  pad_w = 4.5; pad_h = 2.5

  # "auto" (the default) picks the largest square cell that keeps the canvas
  # inside max_w x max_h, so a 20-feature panel gets generous cells and a
  # 300-feature panel shrinks instead of demanding a 46-inch figure. A numeric
  # cell_size in YAML pins it and the canvas grows to fit.
  cell = if(is.numeric(cell_size) && length(cell_size) &&
            all(is.finite(cell_size))){
    rep(cell_size, length.out = 2)
  } else {
    # Fill each axis independently, then pull the two back towards each other
    # so a cell is never more than max_aspect from square. Forcing exact
    # squares on a 600-feature x 30-sample panel would waste most of the
    # canvas; 1.5:1 keeps the cells readable while using the space.
    cw = max(min_cell, min(max_cell, (max_w - pad_w) * 72 / max(n_x, 1)))
    ch = max(min_cell, min(max_cell, (max_h - pad_h) * 72 / max(n_y, 1)))
    if(cw / ch > max_aspect) cw = ch * max_aspect
    if(ch / cw > max_aspect) ch = cw * max_aspect
    c(cw, ch)
  }

  # Cells are absolute units, so the canvas is sized to hold them (72 pt/inch).
  w = n_x * cell[1] / 72 + pad_w
  h = n_y * cell[2] / 72 + pad_h
  list(w = max(10, w) * fig_scale, h = max(6, h) * fig_scale, cell = cell)
}
# Cell size for ONE target drawn into an already-fixed canvas. Every tab in a
# section shares one knitr canvas (one fig.width/height per chunk), so a small
# target must use BIGGER cells to fill it - otherwise it floats in whitespace
# while the largest target sets the size.
.hm_cell_for = function(n_x, n_y, w, h, cfg,
                        min_cell = 2, max_cell = 40,
                        pad_w = 4.5, pad_h = 2.5){
  if(is.numeric(cfg$cell_size) && length(cfg$cell_size) &&
     all(is.finite(cfg$cell_size)))
    return(rep(cfg$cell_size, length.out = 2))   # pinned in YAML

  asp = cfg$cell_aspect %||% 1.5
  cw  = max(min_cell, min(max_cell, (w - pad_w) * 72 / max(n_x, 1)))
  ch  = max(min_cell, min(max_cell, (h - pad_h) * 72 / max(n_y, 1)))
  if(cw / ch > asp) cw = ch * asp
  if(ch / cw > asp) ch = cw * asp
  c(cw, ch)
}

.hm_scope_nrow = function(scope){
  n = vapply(.heatmap_targets(scope), function(tg) nrow(tg$M), integer(1))
  if(length(n)) max(n) else nrow(bio_matrix)
}
```

```{r heatmap-general-precompute, include=FALSE, eval=heatmap_include && heatmap_general_on}
.tg45 = .heatmap_targets(heatmap_general_scope)
.d45  = .hm_fig_dims(.hm_scope_nrow(heatmap_general_scope), ncol(bio_matrix),
                     heatmap_general_cfg$orientation, fig_scale,
                     cell_size  = heatmap_general_cfg$cell_size,
                     max_aspect = heatmap_general_cfg$cell_aspect)
```

```{r heatmap-general, results="asis", eval=heatmap_include && heatmap_general_on, fig.width=(if(exists(".d45")) .d45$w else 10), fig.height=(if(exists(".d45")) .d45$h else 6), out.width="100%"}
if(!length(.tg45))
  cat(sprintf("_No render targets for scope '%s'._\n\n", heatmap_general_scope))
for(tg in .tg45){
  cat(sprintf("\n### %s {-}\n\n", tg$label))
  cat(sprintf("_%d samples × %d features_\n\n", nrow(tg$M), ncol(tg$M)))
  .draw_heatmap(tg$M, tg$smeta, tg$vmeta,
                sprintf("%s of %s — %s", heatmap_general_cfg$scale,
                        heatmap_general_cfg$transform, tg$label),
                cfg = heatmap_general_cfg, canvas = .d45)
  cat("\n\n")
}
```

## 4.6 Heatmap — top-changing per class {.tabset .tabset-fade}

`r if(heatmap_include && !heatmap_topchg_on) "_Skipped — set_ \x60heatmap.top_changing.enabled: True\x60 _in YAML to enable._"`

`r if(heatmap_include && heatmap_topchg_on) sprintf("%s Comparison tabs rank by %s; the Global tab ranks by separation across \x60%s\x60 (no single contrast). Scope: **%s**. Feature labels: %s.", .tc_caption(heatmap_topchg_cfg$group_features_by), heatmap_topchg_rank, heatmap_topchg_cfg$color_samples_by, heatmap_topchg_scope, heatmap_topchg_cfg$feature_labels)`

```{r heatmap-topchg-precompute, include=FALSE, eval=heatmap_include && heatmap_topchg_on}
.tg46 = .heatmap_targets(heatmap_topchg_scope)
.gf46 = heatmap_topchg_cfg$group_features_by
.n_cls46 = if(!is.null(.gf46) && .gf46 %in% colnames(bio_de$variable_meta)){
  v = as.character(bio_de$variable_meta[[.gf46]]); length(unique(v[!is.na(v)]))
} else 1L
.d46 = .hm_fig_dims(.hm_scope_nrow(heatmap_topchg_scope),
                    max(2, min(ncol(bio_matrix), heatmap_topchg_n * .n_cls46)),
                    heatmap_topchg_cfg$orientation, fig_scale,
                    cell_size  = heatmap_topchg_cfg$cell_size,
                    max_aspect = heatmap_topchg_cfg$cell_aspect)
```

```{r heatmap-topchg, results="asis", eval=heatmap_include && heatmap_topchg_on, fig.width=(if(exists(".d46")) .d46$w else 10), fig.height=(if(exists(".d46")) .d46$h else 6), out.width="100%"}
if(!length(.tg46))
  cat(sprintf("_No render targets for scope '%s'._\n\n", heatmap_topchg_scope))
for(tg in .tg46){
  cat(sprintf("\n### %s {-}\n\n", tg$label))
  kf = if(is.null(tg$comp))
         .top_changing_global(colnames(tg$M), tg$M, tg$smeta, tg$vmeta,
                              heatmap_topchg_cfg$color_samples_by,
                              heatmap_topchg_cfg$group_features_by,
                              heatmap_topchg_cfg$transform)
       else
         .top_changing(tg$comp$name, colnames(tg$M), tg$vmeta,
                       group_by = heatmap_topchg_cfg$group_features_by)
  Mk = tg$M[, kf, drop = FALSE]
  rank_desc = if(is.null(tg$comp)) sprintf("%s separation", heatmap_topchg_cfg$color_samples_by) else heatmap_topchg_rank
  cat(sprintf("_%d samples × %d features (top %d per class by %s)_\n\n",
              nrow(Mk), ncol(Mk), heatmap_topchg_n, rank_desc))
  .draw_heatmap(Mk, tg$smeta, tg$vmeta,
                sprintf("%s of %s — %s (top-changing)", heatmap_topchg_cfg$scale,
                        heatmap_topchg_cfg$transform, tg$label),
                cfg = heatmap_topchg_cfg, canvas = .d46)
  cat("\n\n")
}
```

## 4.7 Heatmap — by class {.tabset .tabset-fade}

`r if(heatmap_include && (!heatmap_split_on || is.null(heatmap_subclass_col))) "_Skipped — set_ \x60heatmap.split_by_class.enabled: True\x60 _and a_ \x60subclass_col\x60 _in YAML to enable._"`

`r if(heatmap_include && heatmap_split_on && !is.null(heatmap_subclass_col)) sprintf("Split by \x60%s\x60 (%s). Scope: **%s**. %s Feature labels: %s.", heatmap_subclass_col, if(length(heatmap_subclass_vals)) paste(heatmap_subclass_vals, collapse=", ") else "all values", heatmap_split_scope, if(heatmap_split_multi=="combined") "Selected values combined in one heatmap (grouped by class)." else "One heatmap per value.", heatmap_split_cfg$feature_labels)`

```{r heatmap-split-precompute, include=FALSE, eval=heatmap_include && heatmap_split_on && !is.null(heatmap_subclass_col)}
.tg47 = .heatmap_targets(heatmap_split_scope)
# 4.7 draws only the selected subclass values, so size the canvas to those
# features - not to the whole panel, which would leave the drawn heatmap
# floating in whitespace.
.n_feat47 = local({
  vm = as.data.frame(bio_de$variable_meta)
  if(is.null(heatmap_subclass_col) ||
     !heatmap_subclass_col %in% colnames(vm)) return(ncol(bio_matrix))
  v    = as.character(vm[[heatmap_subclass_col]])
  vals = if(length(heatmap_subclass_vals)) heatmap_subclass_vals
         else unique(v[!is.na(v)])
  n_per = vapply(vals, function(k) sum(v == k, na.rm = TRUE), integer(1))
  # "combined" puts every selected class in one heatmap; "tabs" draws one per
  # class, so the canvas only needs to hold the largest.
  if(identical(heatmap_split_multi, "combined")) max(2L, sum(n_per))
  else                                           max(2L, max(n_per))
})

.d47  = .hm_fig_dims(.hm_scope_nrow(heatmap_split_scope), .n_feat47,
                     heatmap_split_cfg$orientation, fig_scale,
                     cell_size  = heatmap_split_cfg$cell_size,
                     max_aspect = heatmap_split_cfg$cell_aspect)
```

```{r heatmap-split, results="asis", eval=heatmap_include && heatmap_split_on && !is.null(heatmap_subclass_col), fig.width=(if(exists(".d47")) .d47$w else 10), fig.height=(if(exists(".d47")) .d47$h else 6), out.width="100%"}
# Per scope target, split features by `subclass_col`. multi_class == "combined"
# draws all selected values in one heatmap (grouped by the subclass column);
# "tabs" draws one heatmap per value. `subclass_values` (if set) focuses values.
split_col = heatmap_subclass_col
for(tg in .tg47){
  vmeta_t = tg$vmeta
  if(!split_col %in% colnames(vmeta_t)){
    cat(sprintf("\n### %s {-}\n\n_subclass_col '%s' not in feature_metadata — skipped._\n\n",
                tg$label, split_col)); next
  }
  vals = sort(unique(as.character(vmeta_t[[split_col]])))
  vals = vals[!is.na(vals) & nzchar(vals)]
  if(length(heatmap_subclass_vals))
    vals = intersect(vals, as.character(heatmap_subclass_vals))
  if(!length(vals)){
    cat(sprintf("\n### %s {-}\n\n_no values match subclass_values._\n\n", tg$label)); next
  }
  if(heatmap_split_multi == "combined"){
    feats_v = intersect(vmeta_t$Compound[as.character(vmeta_t[[split_col]]) %in% vals],
                        colnames(tg$M))
    if(length(feats_v) < 2){
      cat(sprintf("\n### %s {-}\n\n_too few features._\n\n", tg$label)); next
    }
    cat(sprintf("\n### %s {-}\n\n", tg$label))
    cat(sprintf("_%d samples × %d features across %s: %s_\n\n",
                nrow(tg$M), length(feats_v), split_col, paste(vals, collapse = ", ")))
    .draw_heatmap(tg$M[, feats_v, drop = FALSE], tg$smeta, vmeta_t,
                  sprintf("%s — %s", tg$label, paste(vals, collapse = " / ")),
                  cfg = heatmap_split_cfg, canvas = .d47, group_by = split_col)
    cat("\n\n")
  } else {
    for(v in vals){
      feats_v = intersect(vmeta_t$Compound[as.character(vmeta_t[[split_col]]) == v],
                          colnames(tg$M))
      if(length(feats_v) < 2) next
      cat(sprintf("\n### %s — %s: %s {-}\n\n", tg$label, split_col, v))
      cat(sprintf("_%d samples × %d features_\n\n", nrow(tg$M), length(feats_v)))
      .draw_heatmap(tg$M[, feats_v, drop = FALSE], tg$smeta, vmeta_t,
                    sprintf("%s — %s = %s", tg$label, split_col, v),
                    cfg = heatmap_split_cfg, canvas = .d47)
      cat("\n\n")
    }
  }
}
```

# 5. Ion ratios

User-defined compound ratios (e.g. EpOME/DiHOME as sEH activity proxy). Each
ratio is `numerator / denominator` per sample; denominator ≤ 0 gives NA.
Wilcoxon p-value shown per comparison panel (Kruskal–Wallis for ≥3 groups).
Ratios are **not** carried into heatmap, PCA, volcano, or the comparisons table.

`r if(!ion_ratios_enabled) "_No ion ratios defined — add an_ \x60ion_ratios:\x60 _block with a_ \x60ratios:\x60 _list to_ \x60stats_report\x60 _in the YAML to enable this section._"`

```{r ion-ratios, results="asis", eval=ion_ratios_enabled && !is.null(stats_tables), fig.width=max(8, length(comparisons)*3.5)*fig_scale, fig.height=5.5*fig_scale}
has_comps_ir  = length(comparisons) > 0
has_factor_ir = boxplot_enabled && length(boxplot_factors) > 0

for(ratio_cfg in ion_ratios_list){
  num_col   = as.character(ratio_cfg$num)
  den_col   = as.character(ratio_cfg$den)
  ratio_lbl = as.character(ratio_cfg$label %||% paste0(num_col, " / ", den_col))

  ok_num   = num_col %in% colnames(data_matrix)
  ok_den   = den_col %in% colnames(data_matrix)
  use_tabs = has_comps_ir && has_factor_ir

  if(use_tabs) cat(sprintf("\n\n## %s {.tabset}\n\n", ratio_lbl))
  else         cat(sprintf("\n\n## %s {-}\n\n", ratio_lbl))
  cat(sprintf("_Ratio:_ `%s` / `%s`\n\n", num_col, den_col))

  if(!ok_num || !ok_den){
    if(!ok_num) cat(sprintf("_Numerator `%s` not found in data._\n\n", num_col))
    if(!ok_den) cat(sprintf("_Denominator `%s` not found in data._\n\n", den_col))
    next
  }

  # Values come from runStats()$ion_ratios, so the plots and the ion_ratios
  # sheet of the stats workbook are the same numbers.
  .ir_all  = stats_tables$ion_ratios
  .ir_this = .ir_all[.ir_all$ratio == ratio_lbl & is.na(.ir_all$comparison), ,
                     drop = FALSE]
  ratio_all = stats::setNames(.ir_this$value, .ir_this$sample)

  # ---- By comparison -------------------------------------------------------
  if(has_comps_ir){
    if(use_tabs) cat("\n\n### By comparison {-}\n\n")

    long_rows_ir = list()
    anno_rows_ir = list()

    for(comp in comparisons){
      if(!isTRUE(comp$enabled %||% TRUE)) next
      cond = if(is.null(comp$subset)) list() else as.list(comp$subset)
      cond[[comp$compare$factor]] = comp$compare$levels
      de_sub = tryCatch(subsetDataset(importData, conditions = cond),
                        error = function(e) NULL)
      if(is.null(de_sub) || nrow(de_sub$sample_meta) == 0) next

      sub_rn    = rownames(de_sub$sample_meta)
      sub_ratio = ratio_all[sub_rn]
      sub_grp   = factor(de_sub$sample_meta[[comp$compare$factor]],
                         levels = comp$compare$levels)
      sub_sid   = if(sample_id_head %in% colnames(de_sub$sample_meta))
                    as.character(de_sub$sample_meta[[sample_id_head]]) else sub_rn
      sub_gsf   = if(global_summary_factor %in% colnames(de_sub$sample_meta))
                    as.character(de_sub$sample_meta[[global_summary_factor]]) else NA_character_

      long_rows_ir[[length(long_rows_ir) + 1]] = data.frame(
        .comparison = comp$name,
        .group      = sub_grp,
        .sample_id  = sub_sid,
        ratio_value = sub_ratio,
        .tooltip    = .point_tooltip(sub_sid, sub_gsf, sub_ratio, ylab = "ratio"),
        stringsAsFactors = FALSE
      )

      ngrp = length(comp$compare$levels)
      .ir_comp = .ir_all[.ir_all$ratio == ratio_lbl &
                         !is.na(.ir_all$comparison) &
                         .ir_all$comparison == comp$name, , drop = FALSE]
      pval = if(nrow(.ir_comp)) .ir_comp$p_value[1] else NA_real_

      plab = if(is.na(pval))  "p = NA"    else
             if(pval < 0.001) "p < 0.001" else
             sprintf("p = %.3f", pval)
      if(ngrp > 2) plab = paste0("KW ", plab) else plab = paste0("W ", plab)

      anno_rows_ir[[length(anno_rows_ir) + 1]] = data.frame(
        .comparison = comp$name,
        .group      = comp$compare$levels[length(comp$compare$levels)],
        label       = plab,
        stringsAsFactors = FALSE
      )
    }

    if(length(long_rows_ir) == 0){
      cat("_No valid comparisons for this ratio._\n\n")
    } else {
      long_ir        = dplyr::bind_rows(long_rows_ir)
      long_ir$.group = factor(long_ir$.group, levels = unique(long_ir$.group))
      anno_ir        = dplyr::bind_rows(anno_rows_ir)
      anno_ir$.group = factor(anno_ir$.group, levels = levels(long_ir$.group))

      y_lim = quantile(long_ir$ratio_value, prob = yaxis_quant, na.rm = TRUE) * yaxis_scalar
      y_ann = y_lim * 0.97

      .ir_de = struct::DatasetExperiment(
        data          = data.frame(ratio = long_ir$ratio_value,
                                   row.names = make.unique(long_ir$.sample_id)),
        sample_meta   = local({
          m = data.frame(.group      = long_ir$.group,
                         .comparison = long_ir$.comparison,
                         row.names   = make.unique(long_ir$.sample_id))
          m[[sample_id_head]] = long_ir$.sample_id
          m
        }),
        variable_meta = data.frame(Compound = "ratio",
                                   row.names = "ratio"))

      # The Wilcoxon / Kruskal annotation is report presentation, so it is
      # layered on top of the returned plot rather than baked into
      # plotBoxplot().
      plots_ir = lapply(if(plot_type == "both") c("box", "bar") else plot_type,
        function(ty)
          plotBoxplot(.ir_de, "ratio", ".group", facet_by = ".comparison",
                       type           = ty,
                       error_bar      = plot_error_bar,
                       value_label    = "ratio",
                       sample_id_head = sample_id_head,
                       yaxis_quant    = yaxis_quant,
                       yaxis_scalar   = yaxis_scalar) +
            ggplot2::labs(title = ratio_lbl) +
            ggplot2::geom_text(data = anno_ir,
                               ggplot2::aes(x = .group, label = label),
                               y = y_ann, hjust = 0.5, vjust = 1, size = 3.2,
                               fontface = "italic", color = "grey30",
                               inherit.aes = FALSE))
      names(plots_ir) = if(plot_type == "both") c("box", "bar") else plot_type
      for(pname in names(plots_ir)){
        if(plot_type == "both")
          cat(sprintf('<div style="margin-top:10px;"><b>%s</b></div>\n',
                      if(pname == "box") "Boxplot" else "Barplot"))
        .show_plot(plots_ir[[pname]])
      }
      cat("\n\n")
    }
  }

  # ---- By factor -----------------------------------------------------------
  if(has_factor_ir){
    if(use_tabs) cat("\n\n### By factor {-}\n\n")

    keep_ir  = metadata[[boxplot_head]] %in% boxplot_levels
    ir_meta  = metadata[keep_ir, , drop = FALSE]
    ir_ratio = ratio_all[rownames(ir_meta)]

    for(f in boxplot_factors){
      if(!f %in% colnames(ir_meta)) next
      sid_ir = if(sample_id_head %in% colnames(ir_meta))
                 as.character(ir_meta[[sample_id_head]]) else rownames(ir_meta)
      df_f = data.frame(group     = as.character(ir_meta[[f]]),
                        value     = ir_ratio,
                        sample_id = sid_ir,
                        gsf       = if(global_summary_factor %in% colnames(ir_meta))
                                      as.character(ir_meta[[global_summary_factor]]) else NA_character_,
                        stringsAsFactors = FALSE)
      df_f = df_f[!is.na(df_f$group), ]
      if(nrow(df_f) == 0) next

      df_f$.tooltip = .point_tooltip(df_f$sample_id, df_f$gsf, df_f$value, ylab = "ratio")

      .irf_de = struct::DatasetExperiment(
        data          = data.frame(ratio = df_f$value,
                                   row.names = make.unique(df_f$sample_id)),
        sample_meta   = local({
          m = data.frame(.group = df_f$group,
                         row.names = make.unique(df_f$sample_id))
          m[[sample_id_head]]        = df_f$sample_id
          m[[global_summary_factor]] = df_f$gsf
          m
        }),
        variable_meta = data.frame(Compound = "ratio", row.names = "ratio"))

      plots_f = lapply(if(plot_type == "both") c("box", "bar") else plot_type,
        function(ty)
          plotBoxplot(.irf_de, "ratio", ".group",
                       type           = ty,
                       error_bar      = plot_error_bar,
                       value_label    = "ratio",
                       sample_id_head = sample_id_head,
                       tooltip_factor = global_summary_factor,
                       yaxis_quant    = yaxis_quant,
                       yaxis_scalar   = yaxis_scalar) +
            ggplot2::labs(title = sprintf("%s — by %s", ratio_lbl, f)))
      names(plots_f) = if(plot_type == "both") c("box", "bar") else plot_type
      for(pname in names(plots_f)){
        if(plot_type == "both")
          cat(sprintf('<div style="margin-top:10px;"><b>%s</b></div>\n',
                      if(pname == "box") "Boxplot" else "Barplot"))
        .show_plot(plots_f[[pname]])
      }
      cat("\n\n")
    }
  }
}
```

# 6. Correlations (feature × feature) {.tabset}

Pairwise feature correlations within each subset, one heatmap per
(correlation × subset × method).

> **Long table of all pairs (estimates, statistics, p-values, p_adj) is
> in the `correlations` tab of the stats xlsx.**

## Heatmaps {-}

```{r corr-heatmaps, results="asis", eval=!is.null(stats_tables) && nrow(stats_tables$correlations) > 0, fig.width=8*fig_scale, fig.height=7*fig_scale}
# One heatmap per (correlation × subset × method).
# Optional clustering controlled by `correlations.entries[i].cluster:` in
# YAML; `correlations` (unpacked at the top) is the entries list.
corr_yaml = correlations
get_yaml_opt = function(name, opt, default){
  for(c in corr_yaml) if(identical(c$name, name) && !is.null(c[[opt]])) return(c[[opt]])
  default
}

cdf = stats_tables$correlations
combos = unique(cdf[, c("correlation", "subset", "method")])

for(i in seq_len(nrow(combos))){
  cn = combos$correlation[i]; sb = combos$subset[i]; mt = combos$method[i]
  cluster_yn   = isTRUE(get_yaml_opt(cn, "cluster",      TRUE))
  show_values  = isTRUE(get_yaml_opt(cn, "show_values",  FALSE))

  cat(sprintf("\n### %s — %s — %s {-}\n\n", cn, sb, mt))

  p = plotCorrelation(cdf, if(exists("bio_de")) bio_de else NULL,
                       name        = cn,
                       subset      = sb,
                       method      = mt,
                       group_by    = heatmap_group_features_by,
                       cluster     = cluster_yn,
                       show_values = show_values)
  if(is.null(p)){ cat("_(empty)_

"); next }

  print(p); cat("\n\n")
}
```

# 7. Linear models

For each feature we fit `feature ~ formula` (one `lm()` per feature).
Each row of the table is **one coefficient × feature**.

**Reading the terms** (using a model like `~ Treatment + Sex` with reference
levels `Treatment = Ctrl`, `Sex = Female`):

- **`Intercept (ref: Treatment=Ctrl; Sex=Female)`** — the model's expected mean
  value of the feature for a sample at all reference levels (a Ctrl female).
  Often biologically uninteresting on raw intensity data; mostly a baseline.
- **`Treatment: HDM_Curdlan (vs Ctrl)`** — how much higher (positive estimate)
  or lower (negative) the feature is in HDM_Curdlan vs Ctrl, **after adjusting
  for Sex**. This is the treatment effect controlling for sex imbalance.
- **`Sex: Male (vs Female)`** — how much the feature differs in males vs females,
  **after adjusting for Treatment**.
- **Continuous predictors** (e.g. `~ Treatment + Age`) get a single term —
  the per-unit slope of the feature on that variable.
- **Interactions** (e.g. `~ Treatment * Sex`) add a `Treatment:Sex` term that
  asks "does the treatment effect itself differ between males and females".

**`significant` (p_adj < `r sig_threshold`)**: the coefficient is statistically
distinguishable from zero after Benjamini-Hochberg correction across all
features × terms in this run.

`term_raw` keeps the original `lm()` coefficient name (e.g.
`TreatmentHDM_Curdlan`) for joining with downstream scripts.

> **Per-model coefficient tables (estimate, std_error, statistic, p_value,
> p_adj, significant) are in the `linear_models` tab of the stats xlsx.**

# 8. Session info

```{r session}
tryCatch(sessionInfo(), error = function(e){
  cat("sessionInfo() failed (likely a malformed DESCRIPTION in an installed package):\n")
  cat(conditionMessage(e), "\n\n")
  cat("R version:", R.version$version.string, "\n")
  cat("Platform :", R.version$platform, "\n")
  cat("Loaded packages:\n")
  cat(paste(" ", sort(loadedNamespaces()), collapse = "\n"), "\n")
})
```
